---
title: "01_sac_scenarios"
author: "Carmen Soria"
format: html
editor: visual
---

# Hypothetical species distributions and their SAC

Creation of hypothetical species distributions in a grid and calculating their spatial autocorrelation (SAC) to see how SAC can change in time (occupancy) and with spatial scale.

The outcome of this code are Figure 1 and 2 of the manuscript.

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(here, sf, tidyverse, spdep, gridExtra, rcartocolor)
```

### Directories

```{r}
#| label: directory-variables
#| message: false
#| warning: false


source(file.path(here(), "codes", "00_config_file.R"))
```

### Input variables

```{r}
#| label: input-vars
#| message: false
#| warning: false

## Grain size of the fine coarse grid
cell_side <- 10000 # 1000 m = 10 km

## Color palettes for plots
pal_general <- carto_pal(4, "Safe")
# Occupancy (Figure 1)
pal_occupancy <- rep(pal_general[1:3], 3)
# Grain (Figure 2)
pal_grain <- rep(pal_general, 2)
```

### Loading Presence - Absence matrices

`autocorr_occ`: list of matrices with combinations of different levels of occupancy and aggregation `autocorr_grain`: list of matrices with different presence dispositions at two spatial scales

These matrices of hypothetical presences (1) and absences (0) previously generated manually

```{r}
#| label: load-sac-grids
#| message: false
#| warning: false
#| results: hide

## Figure 1: occupancy and autocorrelation
autocorr_occ <- readRDS(file.path(paths$input_path, "sac_grid_occ.rds"))
## Figure 2: grain and autocorrelation
autocorr_grain <- readRDS(file.path(paths$input_path, "sac_grid_grain.rds"))
```

## Presence/absence grids

Transforming these matrices of presence / absence into spatial grids for which we can calculate SAC

-   Create grid function: first we create an empty grid

```{r}
#| label: function-create-grid
#| message: false
#| warning: false


### Function to generate grid specifying number of squares, grain size and projection. 

## For this hypothetical grid we used a 33N UTM projection (CRS 32633)
create.grid <- function(num_squares, grain_size, crs = 32633) {
  # Define the coordinates for the bottom-left corner of the grid
  x_min <- 0
  y_min <- 0

  # Create a list to store the square polygons
  squares <- vector("list", num_squares^2)

  # Create the grid of squares
  count <- 1
  for (i in 0:(num_squares - 1)) {
    for (j in 0:(num_squares - 1)) {
      x <- x_min + i * grain_size
      y <- y_min + j * grain_size
      squares[[count]] <- st_polygon(list(matrix(c(
        x, y,
        x + grain_size, y,
        x + grain_size, y + grain_size,
        x, y + grain_size,
        x, y
      ), ncol = 2, byrow = TRUE)))
      count <- count + 1
    }
  }

  # Convert to an sf object and assign UTM Zone 33N
  grid_sf <- st_sf(geometry = st_sfc(squares))
  st_crs(grid_sf) <- crs 

  return(grid_sf)
}
```

-   Creating grids

```{r}
#| label: create-grids
#| message: false
#| warning: false
#| results: hide

# Fine grain 16x16 grid (10 km side length)
sf_grid_fine <- create.grid(num_squares = 16, grain_size = cell_side)

# Coarse grain 4x4 grid (40 km side length)
sf_grid_coarse <- create.grid(num_squares = 4, grain_size = cell_side * 4)
```

-   Joining spatial grids with presence/absence matrices

    -   `sf_grid_occ_list`: list of spatial grids for the relationship between OCC and SAC
    -   `sf_grid_grain_fine_list`: list of fine-grain spatial grids for the relationship between grain size and SAC
    -   `sf_grid_grain_coarse_list`: list of coarse-grain spatial grids for the relationship between grain size and SAC
    -   `sf_grid_grain_list`: list of fine and coarse-grain spatial grids for the relationship between grain size and SAC

```{r}
#| label: presence-absence-grids
#| message: false
#| warning: false

# Fig 1. Occupancy and SAC
sf_grid_occ_list <- list()
for(i in 1:length(autocorr_occ)) {
  sf_grid_occ_list[[i]] <- sf_grid_fine
  sf_grid_occ_list[[i]]$presence <- c(autocorr_occ[[i]])
}
names(sf_grid_occ_list) <- names(autocorr_occ)


# Fig 2. Grain size and SAC
# Fig 2.1 Fine grain
autocorr_grain_fine <- autocorr_grain[grep("_fine", names(autocorr_grain))]
sf_grid_grain_fine_list <- list()
for(i in 1:length(autocorr_grain_fine)) {
  sf_grid_grain_fine_list[[i]] <- sf_grid_fine
  sf_grid_grain_fine_list[[i]]$presence <- c(autocorr_grain_fine[[i]])
}
names(sf_grid_grain_fine_list) <- names(autocorr_grain_fine)

# Fig 2.2 Coarse grain
autocorr_grain_coarse <- autocorr_grain[grep("_coarse", names(autocorr_grain))]
sf_grid_grain_coarse_list <- list()
for(i in 1:length(autocorr_grain_coarse)) {
  sf_grid_grain_coarse_list[[i]] <- sf_grid_coarse
  sf_grid_grain_coarse_list[[i]]$presence <- c(autocorr_grain_coarse[[i]])
}
names(sf_grid_grain_coarse_list) <- names(autocorr_grain_coarse)


# Fig. 2 Storing fine and coarse grain grids into a list
sf_grid_grain_list <- c(sf_grid_grain_fine_list, sf_grid_grain_coarse_list)
```

## Calculating Spatial autocorrelation of hypothetical distributions

Join Count metrics calculated: Observed JC, Expected JC, Variance JC, Significance JC, and JC difference (average observed number of JC per cell - average expected number of JC per cell)

-   Function to calculate SAC

```{r}
#| label: function-sac-calc-grids
#| message: false
#| warning: false


calc.jc.list <- function(sf_list, column_sac = "presence") {
  # Empty list to store results
  jc_list <- vector("list", length(sf_list))
  names(jc_list) <- names(sf_list)
  
  # Calculate SAC for all grid arrangements
  for(i in seq_along(sf_list)) {
    sf_object <- sf_list[[i]]
    jc_result <- joincount.test(
      as.factor(sf_object[[column_sac]]),
      nb2listw(poly2nb(sf_object, queen = TRUE), style = "B")
      )
    names(jc_result) <- c("Absence_0", "Presence_1")
    
    # Store results in the list
    jc_list[[i]] <- list(
      name = names(sf_list)[[i]],
      presence_n = sum(sf_object[[column_sac]], na.rm = T),
      jc_statistic_pres = jc_result$Presence_1$estimate[1],
      jc_expectation_pres = jc_result$Presence_1$estimate[2],
      jc_variance_pres = jc_result$Presence_1$estimate[3],
      jc_pvalue_pres = jc_result$Presence_1$p.value,
      jc_zscore_pres = (jc_result$Presence_1$estimate[1] - jc_result$Presence_1$estimate[2]) /
        sqrt(jc_result$Presence_1$estimate[3]),
      jc_diff_pres = (jc_result$Presence_1$estimate[1] /
                        sum(sf_object[[column_sac]], na.rm = TRUE)) - 
        (jc_result$Presence_1$estimate[2] / 
           sum(sf_object[[column_sac]], na.rm = TRUE))
    )
  }
  return(jc_list)
}
```

-   SAC calculation

```{r}
#| label: calculation-sac-grids
#| message: false
#| warning: false


## List of autocorrelation metrics
sac_occ_list <- calc.jc.list(sf_grid_occ_list)
sac_grain_list <- calc.jc.list(sf_grid_grain_list)

## Dataframes 
sac_occ_df <- bind_rows(sac_occ_list, .id = "name")
sac_grain_df <- bind_rows(sac_grain_list, .id = "name")
```

## Grid plots with SAC metrics

-   Function: plotting grids with SAC

```{r}
#| label: function-sac-grid-plots
#| message: false
#| warning: false

### Function to plot grid with SAC data 
plot.grid <- function(sf_list, jc_metrics, i, color_grid) {
  
  out_list <- vector("list", length = length(sf_list))
  names(out_list) <- names(sf_list)
  
  for(i in seq_along(sf_list)) {
    sf_object <- sf_list[[i]]
    jc_result <- jc_metrics[[i]]
    
    # Extract relevant join count metrics
    jc_statistic_pres <- jc_result$jc_statistic_pres
    jc_diff_pres <- jc_result$jc_diff_pres
    jc_zscore_pres <- jc_result$jc_zscore_pres
    
    # Subtitle with JC metrics of presences
    subtitle <- bquote(
      bold("JC")* "=" * .(round(jc_statistic_pres, 2)) * "; " * 
        bold("Z") * "=" * .(round(jc_zscore_pres, 2)) * "; " * 
        bold("JCD") * "=" * .(round(jc_diff_pres, 2)))


    # Plot
    out_list[[i]] <- ggplot(sf_object) +
      geom_sf(aes(fill = factor(presence))) +
      scale_fill_manual(values = c("0" = "transparent", "1" = color_grid[i]), 
                        name = "Presence") +
      theme_minimal() +
      labs(caption = subtitle) +
      theme(
        plot.caption = element_text(size = 9, hjust = 0.5),  
        legend.position = "none",
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(),  
        panel.grid = element_blank(),   
        plot.margin = margin(3, 3, 3, 3, "pt")  
    )
  }
  return(out_list)
}
```

-   Generating grid plots

```{r}
#| label: grid-plots-sac
#| message: false
#| warning: false


### Grid plots with SAC values Figure 1
grid_occ_list <- plot.grid(sf_grid_occ_list, 
                           sac_occ_list, 
                           color_grid = pal_occupancy)

### Grid plots with SAC values Figure 2
grid_grain_list <- plot.grid(sf_grid_grain_list,
                             sac_grain_list,
                             color_grid = pal_grain)
```

## Barplots

Barplots of the SAC metrics of our grids. Two desired outputs:

1.  Fig. 1: SAC in time (aggregation vs. occupancy)
2.  Fig. 2: SAC across grain sizes (presence arrangement vs. grain size)

-   Function to select the adequate scenarios for the barplots

```{r}
#| label: barplot-selection-function
#| message: false
#| warning: false

### Function to filter groups based on type of occupancy or grain
filter.groups <- function(df, id_column, pattern, levels) {
  df %>%
    filter(grepl(pattern, .[[id_column]])) %>%
    mutate(across(all_of(id_column), ~ factor(., levels = levels)))
}
```

#### Figure 1

-   Function for Fig.1 barplots

```{r}
#| label: barplots-fig1-function
#| message: false
#| warning: false

# Function for Figure 1 barplots
plot.bar.jc.fig1 <- function(occ_df, color_bar, y_lim, x_lab) {
  ggplot(occ_df, aes(x = name_label, y = jc_diff_pres, fill = name)) +
    geom_bar(stat = "identity", color = color_bar, width = 0.6) +
    ylim(y_lim[1], y_lim[2]) +
    scale_fill_manual(values = color_bar) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    theme_classic() +
    theme(legend.position = "none", 
          axis.title.x = element_text(size = 9),
          axis.text.x = element_text(size = 9),
          axis.title.y = element_text(size = 9),
          axis.text.y = element_text(size = 9)) + 
    xlab(x_lab) +
    ylab("JCD") 
}
```

-   Define patterns of percentage of occupancy and level of aggregation

```{r}
#| label: barplots-fig1-aggrupations
#| message: false
#| warning: false


# 1. Occupancy patterns
patterns_occ <- c("_25$", "_50$", "_75$")
levels_occ <- list(c("sep_25", "mid_25", "join_25"),
                   c("sep_50", "mid_50", "join_50"),
                   c("sep_75", "mid_75", "join_75"))
col_occ_list <- list(pal_occupancy[1:3], 
                     pal_occupancy[4:6], 
                     pal_occupancy[7:9])


# 2. Aggregation patterns
patterns_agg <- c("^sep_", "^mid_", "^join_")
levels_agg <- list(c("sep_25", "sep_50", "sep_75"),
                   c("mid_25", "mid_50", "mid_75"),
                   c("join_25", "join_50", "join_75"))
col_agg_list <- list(pal_occupancy[c(1,4,7)], 
                     pal_occupancy[c(2,5,8)], 
                     pal_occupancy[c(3,6,9)])
```

-   Plot Fig. 1 barplots

```{r}
#| label: barplots-fig1-plot
#| message: false
#| warning: false


# y-axis limits
y_lim_1 <- c(min(sac_occ_df$jc_diff_pres, na.rm = T), 
             max(sac_occ_df$jc_diff_pres, na.rm = T))


### Occupancy vs SAC diff joincount barplots 

# 1. Occupancy percentage
barplot_jc_occ_perc_list <- list()

for(i in seq_along(patterns_occ)) {
  barplot_df <- filter.groups(sac_occ_df, "name", patterns_occ[[i]], levels_occ[[i]]) %>% 
    mutate(name_label = factor(
      case_when(str_starts(name, "sep") ~ "low",
                str_starts(name, "mid") ~ "mid",
                str_starts(name, "join") ~ "high"),
      levels = c("low", "mid", "high"))
      ) 
  
  barplot_jc_occ_perc_list[[i]] <- plot.bar.jc.fig1(
    barplot_df, col_occ_list[[i]], y_lim = y_lim_1, x_lab = "Aggregation"
    )
  
  names(barplot_jc_occ_perc_list)[i] <- paste0("occupancy", gsub("\\$$", "", patterns_occ[i]))
  
}


# 2. Aggregation level
barplot_jc_occ_agg_list <- list()

for(i in seq_along(patterns_agg)) {
  barplot_df <- filter.groups(sac_occ_df, "name", patterns_agg[[i]], levels_agg[[i]]) %>% 
    mutate(name_label = factor(
      paste0(str_split(name, "_", simplify = T)[,2], "%"),
      levels = c("25%", "50%", "75%"))
      )
  
  barplot_jc_occ_agg_list[[i]] <- plot.bar.jc.fig1(
    barplot_df, col_agg_list[[i]], y_lim = y_lim_1, x_lab = "Occupancy"
    )
  names(barplot_jc_occ_agg_list)[i] <- paste0("aggregation_", gsub("^\\^|_$", "", patterns_agg[i]))
}
```

#### Figure 2

-   Function for Fig. 2 barplots

```{r}
#| label: barplots-fig2-function
#| message: false
#| warning: false


# Function for Figure 2 barplots
plot.bar.jc.grain <- function(occ_df, color_bar, y_lim, x_lab) {
  ggplot(occ_df, aes(x = name_label, y = jc_diff_pres, fill = name, width = bar_width)) +
    geom_bar(stat = "identity", color = color_bar) +
    ylim(y_lim[1], y_lim[2]) +
    scale_fill_manual(values = color_bar) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    theme_classic() +
    theme(legend.position = "none", 
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
          axis.title.y = element_text(size = 9),
          axis.text.y = element_text(size = 9)) + 
    xlab(x_lab) +
    ylab("JCD")
}
```

-   Define patterns of presence disposition and grain size

```{r}
#| label: barplots-fig2-aggrupations
#| message: false
#| warning: false

## Define patterns of disposition and grain size
# 1. Disposition patterns
patterns_disp <- list("^edges_", "^side_", "^center_", "^disperse_")
levels_disp <- list(c("edges_fine", "edges_coarse"),
                    c("side_fine", "side_coarse"),
                    c("center_fine", "center_coarse"),
                    c("disperse_fine", "disperse_coarse"))
col_disp_list <- list(pal_grain[c(1,5)], 
                      pal_grain[c(2,6)], 
                      pal_grain[c(3,7)], 
                      pal_grain[c(4,8)])

# 2. Grain size pattern
patterns_grain <- list("_fine$", "_coarse$")
levels_grain <- list(c("edges_fine", "side_fine", "center_fine", "disperse_fine"),
                     c("edges_coarse", "side_coarse", "center_coarse",
                       "disperse_coarse"))
col_grain_list <- list(pal_grain[c(1:4)], 
                       pal_grain[c(5:8)])
```

-   Plot Fig. 2 barplots

```{r}
#| label: barplots-fig2-plot
#| message: false
#| warning: false


# y-axis limits for bar plots
y_lim_2 <- c(min(sac_grain_df$jc_diff_pres, na.rm = T), 
             max(sac_grain_df$jc_diff_pres, na.rm = T))


### Grain size vs SAC diff joincount barplots 

# 1. Grain disposition
barplot_jc_grain_disp_list <- list()

for(i in seq_along(patterns_disp)) {
  barplot_df <- filter.groups(sac_grain_df, "name", patterns_disp[[i]], levels_disp[[i]]) %>% 
    mutate(bar_width = ifelse(grepl("_coarse$", name), 0.9, 0.4),
       disposition = case_when(str_starts(name, "edges") ~ "a",
                               str_starts(name, "side") ~ "b",
                               str_starts(name, "center") ~ "c",
                               str_starts(name, "disperse") ~ "d"),
       grain = str_split(name, "_", simplify = T)[,2],
       name_label = factor(grain, levels = c("fine", "coarse")))
  
  barplot_jc_grain_disp_list[[i]] <- plot.bar.jc.grain(
    barplot_df, col_disp_list[[i]],  y_lim = y_lim_2, x_lab = "Grain size"
    )
  names(barplot_jc_grain_disp_list)[i] <- paste0("disposition_", gsub("^\\^|_$", "", patterns_disp[i]))
}

```

## Arranged figures

Figures for the manuscript of the different grid dispositions, their respective diff join count and join count p values, as well as the barplots of the diff join count values.

#### Fig. 1: Occupancy vs. SAC

```{r}
#| label: arranged-fig1
#| message: false
#| warning: false


# Barplot occupancy plot
barplot_jc_occ_fig_list <- list(grid_occ_list$sep_25,
                                grid_occ_list$sep_50,
                                grid_occ_list$sep_75,
                                barplot_jc_occ_agg_list$aggregation_sep,
                                grid_occ_list$mid_25,
                                grid_occ_list$mid_50,
                                grid_occ_list$mid_75,
                                barplot_jc_occ_agg_list$aggregation_mid,
                                grid_occ_list$join_25,
                                grid_occ_list$join_50,
                                grid_occ_list$join_75,
                                barplot_jc_occ_agg_list$aggregation_join,
                                barplot_jc_occ_perc_list$occupancy_25,
                                barplot_jc_occ_perc_list$occupancy_50,
                                barplot_jc_occ_perc_list$occupancy_75)
bar_jc_occ_fig <- grid.arrange(grobs = barplot_jc_occ_fig_list, ncol = 4)
```

#### Fig. 2: Grain vs. SAC

```{r}
#| label: arranged-fig2
#| message: false
#| warning: false

# Barplot grain plot
barplot_jc_grain_fig_list <- list(grid_grain_list$edges_fine,
                                  grid_grain_list$side_fine,
                                  grid_grain_list$center_fine,
                                  grid_grain_list$disperse_fine,
                                  grid_grain_list$edges_coarse,
                                  grid_grain_list$side_coarse,
                                  grid_grain_list$center_coarse,
                                  grid_grain_list$disperse_coarse,
                                  barplot_jc_grain_disp_list$disposition_edges,
                                  barplot_jc_grain_disp_list$disposition_side,
                                  barplot_jc_grain_disp_list$disposition_center,
                                  barplot_jc_grain_disp_list$disposition_disperse)
bar_jc_grain_fig <- grid.arrange(grobs = barplot_jc_grain_fig_list, ncol = 4)
```

-   Saving outputs

```{r}
#| label: save-figs
#| message: false
#| warning: false

# Fig. 1
ggsave(file.path(paths$figure_path, "fig1_sacocc1.png"), 
       plot = bar_jc_occ_fig,  width = 7, height = 5, dpi = 600, bg="white")

# Fig. 2
ggsave(file.path(paths$figure_path, "fig2_sacgrain1.png"), 
       plot = bar_jc_grain_fig, width = 8, height = 5, dpi = 600, bg="white")
```
