---
title: "03_data_preparation"
author: "Carmen Soria"
format: html
editor: visual
---

# Data preparation for SAC calculation

Preparation of the BBA data for the calculation of Spatial Autocorrelation

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, here)
```

### Directory variables

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Loading data

```{r}
#| label: load-data
#| message: false
#| warning: false

data_tbl <- readRDS(file.path(paths$input_path, "species_data.rds"))
```

## Data processing

### Excluding Japan BBA3, changing datasetID name, converting to factors and ordering

The third Japanese BBA has a way higher survey effort than the first two. This is evidenced by the addition of 147 novel species that were not in the previous atlases. To ensure compatibility between BBAs we have decided to not consider this atlas replication.

Here we also change the datasetID to region name, convert ID, scale and sampling period to an ordered factor and remove recordFilter (no longer relevant).

```{r}
#| label: data-formatting
#| message: false
#| warning: false


data_tbl1 <- data_tbl %>% 
  filter(!(datasetID == 13 & samplingPeriodID == 3)) %>% 
  mutate(
    datasetID = factor(case_when(
      datasetID == 5 ~ "Czechia",
      datasetID == 26 ~ "Europe",
      datasetID == 6 ~ "New York",
      datasetID == 13 ~ "Japan"
    ), levels = c("Czechia", "Europe", "New York", "Japan"))
  ) %>% 
  mutate(across(c(scalingID, samplingPeriodID), as.factor)) %>% 
  select(-recordFilter)
```

### Species

Identifying and removing observations of hybrid species

```{r}
#| label: removing-hybrids
#| message: false
#| warning: false


# Which species are hybrid and where?
data_tbl1 %>% 
  st_drop_geometry() %>% 
  select(datasetID, scientificName, taxonRank) %>% 
  filter(taxonRank == "hybrid") %>% 
  distinct()


 # Dropping hybrids
data_tbl2 <- data_tbl1 %>% 
  filter(taxonRank != "hybrid") %>% 
  select(-taxonRank)
```

Identifying species that have a different verbatim and scientific name (scientific name follows the HBW/BL v9.1 taxonomy)

```{r}
#| label: nonmatching-names
#| message: false
#| warning: false


# Table with species for which the verbatim name and the BL2024 taxonomy do not match
data_tbl_sp <- data_tbl2 %>% 
  select(verbatimIdentification, scientificName) %>% 
  filter(verbatimIdentification != scientificName) %>% 
  distinct()

# How many cases?
nrow(data_tbl_sp)

# Save mismatch table
write.csv(data_tbl_sp, file.path(paths$output_path, "mismatch_taxonomy.csv"))
```

Keeping only HBW/BL v9.1 taxonomy and removing subspecies (i.e., duplicate rows).

```{r}
#| label: scientific-names-filter
#| message: false
#| warning: false


data_tbl3 <- data_tbl2 %>% 
  select(-verbatimIdentification) %>% 
  distinct()
```

### Grids

#### Cells sampled in all periods

Identifying cells that have been surveyed in all sampling periods for each dataset

```{r}
#| label: identify-cells-sampled-all-periods
#| message: false
#| warning: false

cells_samp_tbl <- data_tbl3 %>% 
  select(datasetID, scalingID, siteID, samplingPeriodID) %>% 
  group_by(datasetID, scalingID, siteID) %>% 
  mutate(nRepetitions = n_distinct(samplingPeriodID)) %>% 
  group_by(datasetID, scalingID) %>% 
  filter(nRepetitions == max(as.numeric(samplingPeriodID))) %>% 
  distinct()

# Double check. Czechia has 3 replications and Europe, Japan and New York have 2
cells_samp_tbl %>%
  select(-samplingPeriodID) %>% 
  group_by(datasetID) %>% 
  distinct(nRepetitions)
```

Filtering the data to keep only those cells

```{r}
#| label: filter-cells-sampled-all-periods
#| message: false
#| warning: false

data_tbl4 <- data_tbl3 %>% 
  right_join(cells_samp_tbl) %>% 
  select(-nRepetitions)
```

#### Spatial scales with at least 30 cells

Identifying sampling scales with 30 cells or more

```{r}
#| label: scales-over-30-cells
#| message: false
#| warning: false

scales_over30_tbl <- data_tbl4 %>% 
  select(datasetID, scalingID, siteID) %>% 
  distinct() %>% 
  group_by(datasetID, scalingID) %>% 
  mutate(nCells = n_distinct(siteID)) %>% 
  select(-siteID) %>% 
  distinct() %>% 
  filter(nCells >= 30)
```

Filtering the data to keep only those spatial scales

```{r}
#| label: filter-scales
#| message: false
#| warning: false

data_tbl5 <- data_tbl4 %>% 
  right_join(scales_over30_tbl) %>% 
  arrange(datasetID, samplingPeriodID, scalingID) %>% 
  select(-nCells)

# Which scales are we keeping?
data_tbl5 %>% 
  select(datasetID, scalingID) %>% 
  arrange(datasetID, scalingID) %>% 
  distinct()
```

### Saving

```{r}
#| label: save-output
#| message: false
#| warning: false

saveRDS(data_tbl5, file.path(paths$output_path, "species_filtered.rds"))
```
