---
title: "04_derived_metrics"
author: "Carmen Soria"
format: html
editor: visual
---

# Calculating derived metrics

Number of species per cell in each combination of region, scale, and period

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, here)
```

### Directory variables

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Loading data

```{r}
#| label: load-data
#| message: false
#| warning: false

data_tbl <- readRDS(file.path(paths$output_path, "species_filtered.rds"))
```

## Richness

Richness in each cell for each combination of study area, period and grain size

```{r}
#| label: calculate-richness-tbl
#| message: false
#| warning: false

# Obtain richness
richness_tbl <- data_tbl %>% 
  group_by(datasetID, scalingID, siteID, startYear, endYear, samplingPeriodID) %>% 
  summarise(richness = n_distinct(scientificName))

# Ensuring the number of rows is consistent
nrow(richness_tbl) == nrow(data_tbl %>% select(datasetID, scalingID, siteID, startYear) %>% unique())
```

Saving richness

```{r}
#| label: save-richness
#| message: false
#| warning: false

saveRDS(richness_tbl, file.path(paths$output_path, "richness_table.rds"))
```

Number of species per atlas region and atlas region x period

```{r}
#| label: total-species-region-period
#| message: false
#| warning: false

## Total number of species per region
total_region_tbl <- data_tbl %>% 
  group_by(datasetID) %>% 
  summarise(richness = n_distinct(scientificName))

## Number of species per atlas region and period
total_period_tbl <- data_tbl %>% 
  group_by(datasetID, startYear, endYear, samplingPeriodID) %>% 
  summarise(richness = n_distinct(scientificName))
```

Saving

```{r}
#| label: save-richness-metrics
#| message: false
#| warning: false


# Richness per region
write.csv(total_region_tbl, file.path(paths$output_path, "richness_region.csv"))

# Richness per region x period
write.csv(total_period_tbl, file.path(paths$output_path, "richness_period.csv"))
```

## Occupancy

Occupancy of each species for each combination of study area, period and grain size

```{r}
#| label: calculate-occupancy
#| message: false
#| warning: false

# Total number of cells 
total_occ_tbl <- data_tbl %>% 
  select(datasetID, scalingID, siteID, area, croppedArea) %>% 
  distinct() %>% 
  group_by(datasetID, scalingID) %>% 
  summarise(totalCells = n_distinct(siteID),
            totalArea = sum(area),
            totalCroppedArea = sum(croppedArea)) %>% 
  ungroup()

# Calculating species level occupancy
occ_tbl <- data_tbl %>% 
  group_by(datasetID, scalingID, startYear, endYear, samplingPeriodID, scientificName) %>% 
  summarise(occCells = n_distinct(siteID),
            occArea = sum(area),
            occCroppedArea = sum(croppedArea)) %>% 
  ungroup() %>% 
  left_join(total_occ_tbl) %>% 
  mutate(percentageOcc = round((occCells / totalCells) * 100, 2))

# Ensuring the number of rows is consistent
nrow(occ_tbl) == nrow(data_tbl %>% select(datasetID, scalingID, startYear, scientificName) %>% unique())
```

Saving occupancy

```{r}
#| label: save_output
#| message: false
#| warning: false

saveRDS(occ_tbl, file.path(paths$output_path, "occupancy_table.rds"))
```

## Area

Area covered by the atlases

```{r}
#| label: calculate-occupancy
#| message: false
#| warning: false

# Total area covered by the atlases
total_area_tbl <- data_tbl %>% 
  select(datasetID, scalingID, siteID, area, croppedArea) %>% 
  filter(scalingID == 1) %>% 
  distinct() %>% 
  group_by(datasetID) %>% 
  summarise(totalArea = sum(area),
            totalCroppedArea = sum(croppedArea)) %>% 
  ungroup()
```
