---
title: "06_sac_species"
author: "Carmen Soria"
format: html
editor: visual
---

# Calculating species Spatial Autocorrelation

### Library

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, sf, spdep, tictoc, here)
```

### Directory variables

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Loading data

```{r}
#| label: load-data
#| message: false
#| warning: false


# Species presence data
data_tbl <- readRDS(file.path(paths$output_path, "species_sf.rds"))

# Occupancy table
occ_tbl <- readRDS(file.path(paths$output_path, "occupancy_table.rds"))
```

## Species SAC

Calculating global SAC metrics for each species for each combination of area of study, time period and grain size. We are calculating two metrics (Global Moran's I and Join count) for a Queen neighbouring scheme (all immediately adjacent cells). I am calculating both Moran's I and Join count but use only Join counts for the final ms. I am using binary (B) weighing scheme for Join count and a row standardised weighting scheme (W) for Moran's I.

### Functions

1.  Subsetting species

Getting an sf object that covers the entire study area (i.e. all cells) in which presence equals 1 when the species is present, and 0 when it is absent

```{r}
#| label: function-subset-species
#| message: false
#| warning: false

### Subset species presence data and convert the rest of the grid into absences
subset.sp.data <- function(unitsp, sp_name) {
  filtered_data <- unitsp %>%
    mutate(scientificName = ifelse(scientificName != sp_name, NA, scientificName)) %>%
    group_by(across(-c(scientificName, geometry))) %>%
    slice(which.max(!is.na(scientificName))) %>%
    mutate(presence = ifelse(!is.na(scientificName), 1, 0))
  
  return(filtered_data)
}
```

2.  Calculating SAC

```{r}
#| label: function-sac-species
#| message: false
#| warning: false


### Calculating the global SAC (join count and moran's I) of a species distribution
calc.gsac.sp <- function(species_sf, jc_style = "B", moran_style = "W"){
  # Assigning an NA in case SAC can't be calculated (i.e., all cells are occupied)
  moran_q <- NA
  jc_q <- NA

  # Create queen contiguity neighbourhoods
  nb_q <- poly2nb(species_sf, queen = TRUE)

  ## Join count calculation
  tryCatch({
    jc_q <- joincount.test(as.factor(species_sf$presence),
                           nb2listw(nb_q, style = jc_style, zero.policy = T))
    names(jc_q) <- c("Absence", "Presence")
  }, error = function(e) {
    jc_q <- NA
  })

  ## Moran's I (queen)
  tryCatch({
    moran_q <- moran.test(species_sf$presence,
                          nb2listw(nb_q, style = moran_style, zero.policy = TRUE))
  }, error = function (e) {
    moran_q <- NA
  })
  
  ## Create results dataframe
  results_df <- data.frame(
    datasetID = unique(species_sf$datasetID),
    startYear = unique(species_sf$startYear),
    endYear = unique(species_sf$endYear),
    samplingPeriodID = unique(species_sf$samplingPeriodID),
    scalingID = unique(species_sf$scalingID),
    scientificName = unique(na.omit(species_sf$scientificName)),
    nPresence = sum(species_sf$presence),
    
    # Join count metrics
    jcStatistic = if(is.list(jc_q)) jc_q$Presence$estimate[1] else NA,
    jcExpectation = if(is.list(jc_q)) jc_q$Presence$estimate[2] else NA,
    jcVariance = if(is.list(jc_q)) jc_q$Presence$estimate[3] else NA,
    jcPval = if(is.list(jc_q)) jc_q$Presence$p.value else NA,
    jcZScore = if(is.list(jc_q)) (jc_q$Presence$estimate[1] - jc_q$Presence$estimate[2]) / sqrt(jc_q$Presence$estimate[3]) else NA,
    jcDifference = if (is.list(jc_q))
      (jc_q$Presence$estimate[1] / sum(species_sf$presence)) -
      (jc_q$Presence$estimate[2] / sum(species_sf$presence)) else NA,
    
    # Moran's I metrics
    moransI = if(is.list(moran_q)) moran_q$estimate[1] else NA,
    moransIPval = if(is.list(moran_q)) moran_q$p.value else NA
  )

  return(results_df)
}
```

### Calculation

```{r}
#| label: calculation-sac-species
#| message: false
#| warning: false


sf_use_s2(FALSE)

tic()
sac_sp_tbl <- data_tbl %>% 
  group_by(datasetID, startYear, endYear, samplingPeriodID, scalingID) %>% 
  group_map(~ {
    # Group metadata (to add later to the output table)
    group_keys <- list(
      datasetID = .y$datasetID,
      samplingPeriodID = .y$samplingPeriodID,
      startYear = .y$startYear,
      endYear = .y$endYear,
      scalingID = .y$scalingID
    )
    
    group_data <- .x
    sp_names <- unique(group_data$scientificName)
    
    map(sp_names, ~ {
      current_sp <- .x
      species_subset <- subset.sp.data(group_data, current_sp) %>% 
        bind_cols(as.data.frame(group_keys))
      calc.gsac.sp(species_subset)
    }) %>% 
      list_rbind()
  }) %>% 
  bind_rows()
toc()

# Removing rownames
rownames(sac_sp_tbl) <- NULL
```

### Formatting output

-   Defining levels again

```{r}
#| label: defining-datasetid-levels
#| message: false
#| warning: false


# Defining levels again
sac_sp_tbl1 <- sac_sp_tbl %>% 
  mutate(
    datasetID = factor(datasetID, levels = c("Czechia", "Europe", 
                                             "New York", "Japan"))) %>% 
  mutate(across(c(scalingID, samplingPeriodID), as.factor))
```

-   Adding occupancy

```{r}
#| label: adding-occupancy
#| message: false
#| warning: false


# Joining SAC and occupancy
sac_sp_tbl2 <- sac_sp_tbl1 %>% 
  full_join(occ_tbl)

# Check: are occ cells in the sac df = to the ones of the occ df?
identical(as.numeric(sac_sp_tbl2$nPresence), 
          as.numeric(sac_sp_tbl2$occCells)) # Yes!

# Remove the duplicated columns
sac_sp_tbl2$nPresence <- NULL
```

-   Saving

```{r}
#| label: save-species-sac
#| message: false
#| warning: false

write.csv(sac_sp_tbl2, file.path(paths$output_path, "sac_species_table.csv"))
```
