---
title: "07_richness_sac"
author: "Carmen Soria"
format: html
editor: visual
---

# Calculating richness autocorrelation

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, sf, spdep, tictoc)
```

### Directories

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Load data

```{r}
#| label: load-data
#| message: false
#| warning: false

data_tbl <- readRDS(file.path(paths$output_path, "richness_sf.rds"))
```

## Richness SAC

Calculating global SAC metrics for richness for each combination of area of study, time period and grain size. We are calculating Global Moran's I using a Queen neighbouring scheme (all immediately adjacent cells) and a row standardised weighting scheme (W).

### Functions

```{r}
#| label: function-richness-sac
#| message: false
#| warning: false


calc.gsac.rich <- function(richness_sf, moran_style = "W") {
  # Assigning an NA in case SAC can't be calculated
  moran_q <- NA
  
  # Create queen contiguity neighbourhoods
  nb_q <- poly2nb(richness_sf, queen = TRUE)
  
  ## Moran's I (queen)
  tryCatch({
    moran_q <- moran.test(richness_sf$richness,
                          nb2listw(nb_q, style = moran_style, 
                                   zero.policy = TRUE))
  }, error = function (e) {
    moran_q <- NA
  })
  
  ## Extract metrics into a dataframe
  result_df <- data.frame(
    moransI = if(is.list(moran_q)) moran_q$estimate[1] else NA,
    moransIPval = if(is.list(moran_q)) moran_q$p.value else NA

  )
}
```

### Calculation

```{r}
#| label: calculation-sac-richness
#| message: false
#| warning: false


sf_use_s2(FALSE)

tic()
sac_rich_tbl <- data_tbl %>%
  group_by(datasetID, scalingID, startYear, endYear, samplingPeriodID) %>% 
  group_map(~ {
    # Group info, to add to the output table
    group_keys <- list(
      datasetID = .y$datasetID,
      samplingPeriodID = .y$samplingPeriodID,
      startYear = .y$startYear,
      endYear = .y$endYear,
      scalingID = .y$scalingID
    )
    
    group_data <- .x
    
    calc.gsac.rich(group_data) %>% 
      bind_cols(as.data.frame(group_keys))
    
  }) %>% 
  bind_rows()
toc()

# Removing rownames
rownames(sac_rich_tbl) <- NULL

# Reordering columns
sac_rich_tbl <- sac_rich_tbl %>% 
  select(datasetID, startYear, endYear, samplingPeriodID, scalingID, moransI, moransIPval)
```

-   Mean and standard deviation of richness SAC

```{r}
#| label: summary-statistics-sac
#| message: false
#| warning: false

# Mean Richness SAC
print(paste("Mean richness SAC = ", round(mean(sac_rich_tbl$moransI), 3)))

# Standard deviation
print(paste("Standard deviation SAC = ", round(sd(sac_rich_tbl$moransI), 3)))
```

-   Saving

```{r}
#| label: save-richness-sac
#| message: false
#| warning: false

write.csv(sac_rich_tbl, file.path(paths$output_path, "sac_richness_table.csv"))
```
