---
title: "08_analyses_sac_time_grain"
author: "Carmen Soria"
format: html
editor: visual
---

# SAC across time and grains

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, sf, viridis, gridExtra, grid, here, broom, rstatix)
```

### Directories

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Variables

```{r}
#| label: variables
#| message: false
#| warning: false


## Color palettes
# Palette atlas regions
region_col <-  c('#4477AA', '#EE6677', '#228833', '#CCBB44')
```

### Load data

```{r}
#| label: load-data
#| message: false
#| warning: false

# Richness
data_richness <- read.csv(file.path(paths$output_path, "sac_richness_table.csv")) %>% 
  dplyr::select(-X) %>% 
  mutate(cellSide = case_when(
    datasetID == "Czechia" ~ scalingID * 11, # Cell side = 11 km
    datasetID == "Europe" ~ scalingID * 50, # Cell side = 50 km
    datasetID == "New York" ~ scalingID * 5, # Cell side = 5 km
    datasetID == "Japan" ~ scalingID * 20, # Cell side = 50 km
  ))

# Species
data_species <- read.csv(file.path(paths$output_path, "sac_species_table.csv")) %>%
  dplyr::select(-X) %>% 
  mutate(cellSide = case_when(
    datasetID == "Czechia" ~ scalingID * 11, # Cell side = 11 km
    datasetID == "Europe" ~ scalingID * 50, # Cell side = 50 km
    datasetID == "New York" ~ scalingID * 5, # Cell side = 5 km
    datasetID == "Japan" ~ scalingID * 20, # Cell side = 50 km
  ))

# Dataset IDs (same for richness and species)
datasetIDs <- unique(data_richness$datasetID)

# Grid attributes (for the sensitivity test using real average cell side length)
grid_attr <- read.csv(file.path(paths$output_path, "grid_attributes_tbl.csv"))
```

## 1. How autocorrelated are richness and species distributions?

### Species richness

Which combinations of study area, sampling period, and spatial grain have significantly autocorrelated richness?

```{r}
#| label: richness-significant-sac
#| message: false
#| warning: false

rich_significance_tbl <- data_richness %>% 
  dplyr::select(datasetID, startYear, endYear, samplingPeriodID, scalingID, cellSide, moransI, moransIPval) %>% 
  mutate(
    period = paste(startYear, endYear, sep = "-"),
    moransI = round(moransI, 3),
    moransIPval = ifelse(moransIPval < 0.01, "<0.01", 
                         as.character(round(moransIPval, 3))))

# Mean SAC per region
rich_significance_tbl %>% 
  group_by(datasetID) %>% 
  summarise(mean_moransI = mean(moransI),
            sd_moransI = sd(moransI))
```

-   Saving

```{r}
#| label: save-richness-significant-sac
#| message: false
#| warning: false

# Saving
write.csv(rich_significance_tbl, file.path(paths$output_path, "richness_sac_pval.csv"))
```

### Species distributions

How many species have significantly autocorrelated distributions?

The number of species for which we can calculate the join count data will vary between study areas, time periods and grain sizes due to: (a) number of sampled species, (b) number of sampled species for which we can calculate SAC, and (c) the inclusion of cells that were not considered at the smaller grain size.

SAC cannot be calculated for species in all grid cells, we need both presences and absences.

```{r}
#| label: distribution-significant-sac
#| message: false
#| warning: false


# How many species with significant positive and negative SAC? 
species_significance_tbl <- data_species %>% 
  dplyr::select(datasetID, scientificName, samplingPeriodID, scalingID, cellSide,jcZScore, jcPval, jcDifference) %>% 
  group_by(datasetID, samplingPeriodID, scalingID) %>%
  summarise(totalSpecies = n_distinct(scientificName),
            speciesSAC = n_distinct(scientificName[!is.na(jcPval)]),
            percSpeciesSAC = round((speciesSAC/totalSpecies) * 100, 2),
            speciesSignificantPosSAC = n_distinct(scientificName[jcZScore >= 1.96 & 
                                                                !is.na(jcPval)]),
            speciesSignificantNegSAC = n_distinct(scientificName[jcZScore <= -1.96 & 
                                                                !is.na(jcPval)]),
            percSpeciesSignificantPosSAC = round((speciesSignificantPosSAC / 
                                                 speciesSAC) * 100, 2),
            percSpeciesSignificantNegSAC = round((speciesSignificantNegSAC / 
                                                 speciesSAC) * 100, 2)
            )

# Mean per region
species_significance_tbl %>% 
  group_by(datasetID) %>% 
  summarise(mean = mean(percSpeciesSignificantPosSAC),
            sd = sd(percSpeciesSAC))
```

-   Saving

```{r}
#| label: save-species-significant-sac
#| message: false
#| warning: false

# Saving
write.csv(species_significance_tbl, file.path(paths$output_path, "species_sac_pval.csv"))
```

-   Overall mean and median % of species with significant positive SAC

```{r}
#| label: mean-species-significant-sac
#| message: false
#| warning: false

# Mean
paste("mean", round(mean(species_significance_tbl$percSpeciesSignificantPosSAC, 
                         na.rm = T), 2))

# SD
paste("sd", round(sd(species_significance_tbl$percSpeciesSignificantPosSAC, 
                         na.rm = T), 2))

# Median
paste("median", round(median(species_significance_tbl$percSpeciesSignificantPosSAC, 
                             na.rm = T), 2))
```

## 2: Empirical change of SAC over time

Analyzing how SAC changes across time for diversity (species richness) and distributions (species distributions) divided by grain. For species, we also test whether differences in SAC between periods are significant.

### Species richness

#### Temporal trend plot

-   Function for plots (Figure 4a)

```{r}
#| label: function-richness-sac-time-plot
#| message: false
#| warning: false

# Plot function
create.p4rich <- function(dataset_name, data = data_richness) {
  
  change_sac_rich_df <- data %>% 
    filter(datasetID == dataset_name & !is.na(moransI)) %>% 
    mutate(significanceFill = moransIPval < 0.05)
  
  scaling_levels <- n_distinct(change_sac_rich_df$cellSide)
  
  p4_rich <- ggplot(data = change_sac_rich_df,
                    aes(x = startYear, y = moransI,
                        color = as.factor(cellSide),
                        fill = as.factor(ifelse(significanceFill, cellSide, NA)),
                        group = as.factor(cellSide))) +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
    geom_line(linewidth = 0.75) +
    geom_point(size = 2, shape = 21) +
    scale_color_viridis_d(name = "Cell size") +
    scale_x_continuous(breaks = unique(change_sac_rich_df$startYear)) +
    scale_y_continuous(limits = c(-0.3, 0.9), breaks = seq(-0.3, 0.9, by = 0.3)) +
    scale_fill_manual(values = viridis(scaling_levels), 
                      name = "p-value", na.value = "grey50") +
    labs(x = "Start year", y = "Moran's I") +
    theme_classic() +
    theme(legend.position = "none",
          axis.title.x = element_text(size = 10),
          axis.text.x = element_text(size = 9, angle = 45, vjust = 1, hjust = 1),
          axis.title.y = element_text(size = 10),
          axis.text.y = element_text(size = 9))
  
  return(p4_rich)
}
```

-   Obtaining plots

```{r}
#| label: richness-sac-time-plot
#| message: false
#| warning: false

# Plots
time_p4rich_list <- purrr::map(datasetIDs, create.p4rich)
# Naming them
time_p4rich_list <- setNames(time_p4rich_list , datasetIDs)
```

### Species distributions

-   Filtering to keep only species that are in all time periods

```{r}
#| label: temporal-filter-species
#| message: false
#| warning: false

sp_all_periods_df <- data_species %>%
  filter(!is.na(jcDifference)) %>%
  group_by(datasetID, cellSide, scientificName) %>%
  mutate(n_periods = n()) %>%
  ungroup() %>%
  group_by(datasetID) %>%
  filter(n_periods == max(n_periods)) %>% 
  ungroup()
```

#### Count of increasing, decreasing and no change in SAC

How many species are increasing, decreasing and showing no change in SAC over time?

```{r}
#| label: n-species-increasing-decreasing-nochange
#| message: false
#| warning: false

sp_time_count_df <- sp_all_periods_df %>% 
  group_by(datasetID, cellSide, scientificName) %>%
  summarise(
    firstJCD = jcDifference[which.min(samplingPeriodID)],
    lastJCD  = jcDifference[which.max(samplingPeriodID)],
    .groups = "drop"
  ) %>%
  mutate(changeDirection = case_when(
    firstJCD > lastJCD ~ "decrease",
    firstJCD < lastJCD ~ "increase",
    TRUE               ~ "nochange"
  )) %>%
  group_by(datasetID, cellSide, changeDirection) %>%
  summarise(count = n(), .groups = "drop") %>% 
  pivot_wider(names_from = changeDirection, values_from = count, values_fill = 0) %>% 
  rowwise() %>% 
  mutate(n_species = sum(decrease + increase + nochange))

# overall percentage of species increasing over time
paste0("temporal increase: ", round(sum(sp_time_count_df$increase) / sum(sp_time_count_df$n_species) * 100, 2), "% of species")

# overall percentage of species decreasing over time
paste0("temporal decrease: ", round(sum(sp_time_count_df$decrease) / sum(sp_time_count_df$n_species) * 100, 2), "% of species")

# overall percentage of species no changing
paste0("temporal no change: ", round(sum(sp_time_count_df$nochange) / sum(sp_time_count_df$n_species) * 100, 2), "% of species")
```

-   Save

```{r}
#| label: save-nspecies-changes
#| message: false
#| warning: false

write.csv(sp_time_count_df, file.path(paths$output_path, "count_time_sac_sp.csv"))
```

#### Species-level temporal SAC slopes

-   Linear model per dataset x cell size x species

```{r}
#| label: species-temporal-slopes
#| message: false
#| warning: false

sp_time_slopes_df <- sp_all_periods_df %>% 
  group_by(datasetID, cellSide, scientificName) %>% 
  arrange(startYear, .by_group = TRUE) %>% 
  group_modify(~ {
    mod <- lm(jcDifference ~ startYear, data = .x)
    slope_data <- tidy(mod) %>% filter(term == "startYear")
    r2_data <- glance(mod) %>% select(r.squared)
    bind_cols(slope_data, r2_data) %>%
      mutate(n_replicates = nrow(.x))
  }) %>% 
  ungroup() %>% 
  rename(
    slope = estimate,
    se_slope = std.error,
    r2 = r.squared
  )
```

-   Percentage of species with significant slopes in Czechia

```{r}
#| label: species-significant-temporal-cz
#| message: false
#| warning: false

sp_time_slopes_df %>% 
  filter(datasetID == "Czechia") %>% 
  group_by(cellSide) %>% 
  summarise(n_signif = sum(p.value <= 0.05) / n_distinct(scientificName))
```

-   Mean slopes and standard deviations per region and time

```{r}
#| label: mean-temporal-slopes-tbl
#| message: false
#| warning: false

slopes_time_df <- sp_time_slopes_df %>% 
  group_by(datasetID, cellSide) %>% 
  summarise(mean_slope = mean(slope, na.rm = T),
            max_slope = max(slope, na.rm = T),
            min_slope = min(slope, na.rm = T),
            se_slope = sd(slope, na.rm = T),
            mean_r2 = mean(r2))
```

-   Plot mean slopes

```{r}
#| label: mean-temporal-slopes-plot
#| message: false
#| warning: false

ps1_slopes <- ggplot(data = slopes_time_df, 
                      aes(x = cellSide, y = mean_slope, 
                          color = datasetID, group = datasetID)) + 
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  scale_x_log10() +
  ylim(-0.015, 0.015) +
  scale_color_manual(values = region_col, name = "Dataset") +
  labs(x = "Grid resolution (km)", y = "JCD trend") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.position.inside = c(0.93, 0.9),
        legend.spacing.y = unit(0.1, "cm"),
        legend.key.size = unit(0.2, "cm"),
        legend.margin = margin(0, 0, 0, 0),
        plot.margin = margin(t = 5, r = 15, b = 5, l = 5),
        axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 10),
        axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 10))

plot(ps1_slopes)
```

-   Save individual and mean slopes

```{r}
#| label: save-temporal-slopes
#| message: false
#| warning: false

# Slope species df (all)
write.csv(sp_time_slopes_df, file.path(paths$output_path, "slopes_time_sac_sp_all.csv"))

# Slope species df (mean)
write.csv(slopes_time_df, file.path(paths$output_path, "slopes_time_sac_sp_mean.csv"))

# Plot
ggsave(plot = ps1_slopes, width = 5, height = 6, dpi = 300, 
       file.path(paths$figure_path,
                 "ps1_sac_time.png"))
```

#### Temporal trends plot

-   Function temporal trend plot (Figure 4b)

```{r}
#| label: function-sp-sac-time-plot
#| message: false
#| warning: false

# Temporal trend plot
create.p4sp <- function(dataset_name, slopes = sp_time_slopes_df, data = sp_all_periods_df) {
  
  # Filter dataset
  dataset_df <- data %>% 
    filter(datasetID == dataset_name)
  
  # Years info
  year_range <- sort(unique(dataset_df$startYear))
  year0 <- min(year_range)
  
  # Scale info
  scaling_levels <- n_distinct(dataset_df$cellSide)
  
  # Species level intercepts (initial jcd)
  sp_start <- dataset_df %>% 
    dplyr::select(cellSide, scientificName, startYear, jcDifference) %>% 
    group_by(cellSide, scientificName) %>% 
    mutate(firstJCD = jcDifference[which.min(startYear)]) %>% 
    dplyr::select(-startYear, -jcDifference) %>% 
    unique()
  
  # Add to slopes to get regression lines
  sp_reg <- sp_time_slopes_df %>% 
    filter(datasetID == dataset_name) %>% 
    left_join(sp_start, by = c("cellSide", "scientificName")) %>% 
    rowwise() %>% 
    mutate(startYear = list(year_range),
           fitted_reg = list(firstJCD + slope * (year_range - year0))) %>% 
    unnest(c(startYear, fitted_reg))
  
  # Mean line per grain size
  mean_reg <- sp_reg %>% 
    group_by(cellSide, startYear) %>% 
    summarise(fitted_mean = mean(fitted_reg, na.rm = TRUE))
  
  # Max per facet
  y_limits <- sp_reg %>%
    group_by(cellSide) %>%
    summarise(ymax = max(fitted_reg, na.rm = TRUE),
              ymin = min(fitted_reg, na.rm = TRUE))

  
  # Plot
  p4_sp <- ggplot() + 
    geom_line(data = sp_reg, aes(x = startYear, y = fitted_reg, group = scientificName), colour = "black", alpha = 0.05) +
    geom_line(data = mean_reg, aes(x = startYear, y = fitted_mean, color = as.factor(cellSide)), size = 0.75) +
    facet_wrap(~cellSide, ncol = scaling_levels, 
               labeller = labeller(cellSide = ~paste0(.x, " km"))) +
    scale_color_viridis_d(name = "Grid resolution") +
    scale_x_continuous(breaks = unique(sp_reg$startYear)) +
    scale_y_continuous(breaks = pretty(sp_reg$fitted_reg, n = 3)) +
    labs(x = "Start year", y = "JCD") +
    theme_classic() + 
    theme(legend.position = "none",
          axis.title.x = element_text(size = 10),
          axis.text.x = element_text(size = 9, angle = 45, vjust = 1, hjust = 1),
          axis.title.y = element_text(size = 10),
          axis.text.y = element_text(size = 9),
          strip.text = element_text(size = 9),
          panel.spacing = unit(0.5, "lines"))
  
  return(p4_sp)
}
```

-   Plotting

```{r}
#| label: sp-sac-time-plot
#| message: false
#| warning: false

time_p4sp_list  <- purrr::map(datasetIDs, create.p4sp)
# Naming them
time_p4sp_list <- setNames(time_p4sp_list, datasetIDs)
```

#### Statistical analysis

Step 1: look at distributions - they don't look very normal

```{r}
#| label: sp-sac-distributions
#| message: false
#| warning: false

## Histograms of the distributions of JCD for all combinations of region x grain size x time
hist_plots <- ggplot(data = sp_all_periods_df, aes(x = jcDifference)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  facet_grid(datasetID ~ scalingID, scales = "free_y") + 
  labs(
    x = "JCD",
    y = "Count",
    title = "Distribution JCD by dataset and scaling"
  )


boxplot_plots <- ggplot(data = sp_all_periods_df, aes(x = as.factor(samplingPeriodID), y = jcDifference, color = as.factor(samplingPeriodID))) +
  geom_boxplot() + 
  facet_grid(datasetID ~ scalingID) +
  labs(
    x = "start year",
    y = "JCD",
    title = "Boxplots JCD per year, dataset and scaling"
  ) + 
  theme(legend.position = "none")

grid.arrange(hist_plots, boxplot_plots, ncol=2)
```

Step 2.1: Paired samples Wilcoxon test (non-parametric version of the paired t-test) for Europe, New York and Japan (two temporal repetitions)

```{r}
#| label: wilcox-sp-sac-time
#| message: false
#| warning: false

wilc_test <- sp_all_periods_df %>% 
  filter(n_periods == 2) %>% 
  group_by(datasetID, scalingID, scientificName) %>% 
  summarise(val_t1 = jcDifference[samplingPeriodID == 1], 
              val_t2 = jcDifference[samplingPeriodID == 2]) %>% 
  group_by(datasetID, scalingID) %>% 
  summarise({ 
    df <- cur_data() 
    model <- wilcox.test(df$val_t2, df$val_t1, paired = TRUE, alternative = "less") # We assume it's becoming less autocorrelated 
    model_res <- tidy(model) %>% 
      mutate(method = model$method, 
             signif = case_when(
               p.value > 0.05 ~ "ns",
               p.value <= 0.05 & p.value > 0.01 ~ "*",
               p.value <= 0.01 & p.value > 0.001 ~ "**",
               p.value <= 0.001 ~ "***"))
    df_long <- df %>% 
      pivot_longer(cols = starts_with("val_t"), 
                   names_to = "period", 
                   names_prefix = "val_", 
                   values_to = "value") 
    eff <- wilcox_effsize(df_long, 
                          value ~ period, 
                          paired = TRUE) 
    bind_cols(model_res, eff) }) 

```

Step 2.2: Friedman test (non-parametric version of repeated measures ANOVA) with post-hoc wilcoxon pairwise tests

```{r}
#| label: friedman-sp-sac-time
#| message: false
#| warning: false

fried_posthoc <- sp_all_periods_df %>% 
  filter(n_periods == 3) %>% 
  group_by(datasetID, scalingID, scientificName) %>% 
  summarise(
    val_t1 = jcDifference[samplingPeriodID == 1],
    val_t2 = jcDifference[samplingPeriodID == 2],
    val_t3 = jcDifference[samplingPeriodID == 3],
    .groups = "drop"
  ) %>% 
  group_by(datasetID, scalingID) %>% 
  group_modify(~{
    
    df <- .x %>% ungroup()
    
    # Pivot to long format
    df_long <- df %>%
      pivot_longer(
        cols = starts_with("val_t"),
        names_to = "period",
        names_prefix = "val_",
        values_to = "value"
      )
    
    # Global Friedman test
    fried_res <- rstatix::friedman_test(df_long, value ~ period | scientificName)
    eff_res <- rstatix::friedman_effsize(df_long, value ~ period | scientificName)
    
    # Posthoc Wilcoxon for all pairs
    posthoc_res <- df_long %>% 
      pairwise_wilcox_test(
        value ~ period,
        paired = TRUE,
        p.adjust.method = "BH"
      ) %>% 
      mutate(
        effsize = map2_dbl(group1, group2, ~{
          pair_df <- df_long %>% filter(period %in% c(.x, .y))
          wilcox_effsize(pair_df, value ~ period, paired = TRUE)$effsize
        }),
        signif = case_when(
          p.adj > 0.05 ~ "ns",
          p.adj <= 0.05 & p.adj > 0.01 ~ "*",
          p.adj <= 0.01 & p.adj > 0.001 ~ "**",
          p.adj <= 0.001 ~ "***"
        ),
        friedman_statistic = fried_res$statistic,
        friedman_df = fried_res$df,
        friedman_p = fried_res$p,
        friedman_effsize = eff_res$effsize,
        friedman_magnitude = eff_res$magnitude
      )
    
    posthoc_res
  }) %>% ungroup()

fried_posthoc
```

### Richness and species temporal trends plot

-   Arranging the richness and species plots

```{r}
#| label: all-sac-time-plot
#| message: false
#| warning: false

# Arrange the plots and labels into a grid
plot_time_fig <- grid.arrange(
  arrangeGrob(time_p4rich_list$Czechia, 
              time_p4sp_list$Czechia, 
              widths = c(1, 3)),
  arrangeGrob(time_p4rich_list$Europe, 
              time_p4sp_list$Europe,
              widths = c(1, 3)),
  arrangeGrob(time_p4rich_list$`New York`,
              time_p4sp_list$`New York`, 
              widths = c(1, 3)),
  arrangeGrob(time_p4rich_list$Japan, 
              time_p4sp_list$Japan, 
              widths = c(1, 3)),
  nrow = 4
)
```

-   Saving

```{r}
#| label: save-all-sac-time-plot
#| message: false
#| warning: false

ggsave(plot = plot_time_fig, dpi = 300, width = 6, height = 7, file.path(paths$figure_path, "p4_sac_time.png"))
```

## 3: Empirical change of SAC across grains

Analyzing how SAC changes across time for diversity (species richness) and distributions (species distributions) divided by grain.

### Species richness

#### Scaling plots

-   Function

```{r}
#| label: function-richness-sac-grain-plot
#| message: false
#| warning: false


# Plot function
create.p5rich <- function(dataset_name, data = data_richness) {
  
  # Richness moran's I data
  grain_sac_rich_df <- data %>% 
    filter(datasetID == dataset_name & !is.na(moransI)) %>% 
    mutate(startYear = as.factor(startYear))
  
  
  # Plot for Richness Moran's I
  p5_rich <- ggplot(data = grain_sac_rich_df,
                    aes(x = cellSide,
                        y = moransI, 
                        color = as.factor(startYear),
                        fill = as.factor(startYear),
                        group = startYear)) +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
    geom_line(size = 0.75) +
    geom_point(size = 2, shape = 21) +
    scale_x_log10(breaks = unique(grain_sac_rich_df$cellSide),
                  labels = unique(grain_sac_rich_df$cellSide)) +
    scale_color_viridis_d(name = "Start year") +
    scale_fill_viridis_d(name = "Start year") +
    ylim(-0.2, 0.9) +
    labs(x = "Cell side (km)", y = "Moran's I") +
    theme_classic() +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 8, angle = 45, vjust = 1, hjust = 1),
          axis.title.y = element_text(size = 9),
          axis.text.y = element_text(size = 8))
  
  return(p5_rich)
}

```

-   Plotting

```{r}
#| label: richness-sac-grain-plot
#| message: false
#| warning: false

# Plots
grain_p5rich_list <- purrr::map(datasetIDs, create.p5rich)
# Naming them
grain_p5rich_list <- setNames(grain_p5rich_list, datasetIDs)
```

#### Linear models

-   Modelling richness SAC (Moran's I) as a function of grain size

```{r}
#| label: richness-sac-grain-lm
#| message: false
#| warning: false

rich_lm_grain <- data_richness %>% 
  group_by(datasetID, samplingPeriodID) %>% 
  arrange(scalingID, .by_group = TRUE) %>% 
  group_modify(~ {
    mod <- lm(moransI ~ log(cellSide), data = .x)
    slope_data <- tidy(mod) %>% filter(term == "log(cellSide)")
    r2_data <- glance(mod) %>% select(r.squared)
    bind_cols(slope_data, r2_data) %>% 
      mutate(n_grains = nrow(.x))
  }) %>% 
  ungroup() %>% 
  mutate(signif = case_when(
               p.value > 0.05 ~ "ns",
               p.value <= 0.05 & p.value > 0.01 ~ "*",
               p.value <= 0.01 & p.value > 0.001 ~ "**",
               p.value <= 0.001 ~ "***"))
```

-   Save richness lm output

```{r}
#| label: save-richness-sac-grain-lm
#| message: false
#| warning: false

# Save median JCD grain table
write.csv(rich_lm_grain, file.path(paths$output_path, "lm_grain_rich.csv"))
```

-   Sensitivity test (real grain sizes): because the cell sides we are indicating are approximated (not the real ones), we conducted a sensitivity test to see whether these differences are big

```{r}
#| label: richness-sac-grain-lm-sensitivity-test
#| message: false
#| warning: false

rich_lm_grain_v2 <- data_richness %>% 
  left_join(., grid_attr, by = c("datasetID", "scalingID")) %>% 
  group_by(datasetID, samplingPeriodID) %>% 
  arrange(scalingID, .by_group = TRUE) %>% 
  group_modify(~ {
    mod <- lm(moransI ~ log(meanCellSide), data = .x)
    slope_data <- tidy(mod) %>% filter(term == "log(meanCellSide)")
    r2_data <- glance(mod) %>% select(r.squared)
    bind_cols(slope_data, r2_data) %>% 
      mutate(n_grains = nrow(.x))
  }) %>% 
  ungroup() %>% 
  mutate(signif = case_when(
               p.value > 0.05 ~ "ns",
               p.value <= 0.05 & p.value > 0.01 ~ "*",
               p.value <= 0.01 & p.value > 0.001 ~ "**",
               p.value <= 0.001 ~ "***"))
```

### Species distributions

#### Linear models

-   Model considering all species (just filtering out the ones that have no SAC value)

```{r}
#| label: sp-sac-grain-lm
#| message: false
#| warning: false
species_lm_grain <- data_species %>% 
  filter(!is.na(jcDifference)) %>% 
  group_by(datasetID, startYear) %>% 
  group_map(~ { 
    mod <- lm(jcDifference ~ log(cellSide), data = .x) 
    res <- bind_cols(tidy(mod) %>% 
                        filter(term == "log(cellSide)"), 
                      glance(mod) %>% select(r.squared) ) %>% 
      rename( slope = estimate, se_slope = std.error, r2 = r.squared, t_value = statistic ) # add grouping keys from .y 
    cbind(.y, res) }) %>% 
  bind_rows() %>% 
  ungroup() %>% 
  mutate(signif = case_when( p.value > 0.05 ~ "ns", p.value <= 0.05 & p.value > 0.01 ~ "*", p.value <= 0.01 & p.value > 0.001 ~ "**", p.value <= 0.001 ~ "***"))
```

-   Saving

```{r}
#| label: save-sp-sac-grain-lm
#| message: false
#| warning: false

write.csv(species_lm_grain, file.path(paths$output_path, "lm_grain_sp.csv"))
```

-   Sensitivity

#### Plots

-   Function for plots

```{r}
#| label: function-distributions-sac-grain-plot
#| message: false
#| warning: false


# Plot function
create.p5sp <- function(dataset_name, data = data_species, data_slopes = species_lm_grain) {
  
  # Filter and prepare the main plotting data
  grain_sac_sp_df <- data %>% 
    filter(datasetID == dataset_name,
           !is.na(jcDifference), 
           is.finite(jcDifference)) %>% 
    mutate(period = paste(startYear, endYear, sep = "-"))
  
  # Identify which startYears are significant in the slope data
  signif_years <- data_slopes %>%
    filter(datasetID == dataset_name, p.value <= 0.05) %>%
    pull(startYear)
  
  # Base plot (boxplots of JCD by grain size)
  p5_sp <- ggplot(grain_sac_sp_df, 
                  aes(x = cellSide, y = jcDifference)) + 
    geom_boxplot(aes(group = factor(cellSide), fill = factor(startYear)),
                 outlier.size = 1, linewidth = 0.5) +
    facet_wrap(~period, scales = "free") + 
    scale_x_log10(breaks = unique(grain_sac_sp_df$cellSide),
                  labels = unique(grain_sac_sp_df$cellSide)) +
    scale_fill_viridis_d(name = "Start year") +
    theme_classic() +
    labs(x = "Cell side (km)", y = "JCD") +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = 8, angle = 45, vjust = 1, hjust = 1),
          axis.title.y = element_text(size = 9),
          axis.text.y = element_text(size = 8),
          strip.text = element_text(size = 8),
          strip.background = element_blank(),
          panel.spacing = unit(0.5, "lines"))
  
  # Add regression lines only for significant start years
  p5_sp <- p5_sp +
    geom_smooth(
      data = grain_sac_sp_df %>% filter(startYear %in% signif_years),
      aes(x = cellSide, y = jcDifference, group = period),
      method = "lm", formula = y ~ log(x),
      se = FALSE, color = "#CD4F39", linewidth = 1
    )
  
  return(p5_sp)
}
```

-   Plotting

```{r}
#| label: distributions-sac-grain-plot
#| message: false
#| warning: false


# Plots
grain_p5sp_list <- purrr::map(datasetIDs, create.p5sp)
# Naming them
grain_p5sp_list <- setNames(grain_p5sp_list, datasetIDs)
```

#### Linear models

-   Sensitivity test (real grain sizes): because the cell sides we are indicating are approximated (not the real ones), we conducted a sensitivity test to see whether these differences are big

```{r}
#| label: sp-sac-grain-lm-sensitivity-test
#| message: false
#| warning: false


species_lm_grain_v2 <- data_species %>% 
  left_join(., grid_attr, by = c("datasetID", "scalingID")) %>% 
  filter(!is.na(jcDifference)) %>% 
  group_by(datasetID, startYear) %>%
  group_map(~ {
    mod <- lm(jcDifference ~ log(medianArea), data = .x)
    
    res <- bind_cols(
      tidy(mod) %>% filter(term == "log(medianArea)"),
      glance(mod) %>% select(r.squared)
    ) %>%
      rename(
        slope = estimate,
        se_slope = std.error,
        r2 = r.squared,
        t_value = statistic
      )
    
    # add grouping keys from .y
    cbind(.y, res)
  }) %>%
  bind_rows() %>% 
  ungroup() %>% 
  mutate(signif = case_when( p.value > 0.05 ~ "ns", p.value <= 0.05 & p.value > 0.01 ~ "*", p.value <= 0.01 & p.value > 0.001 ~ "**", p.value <= 0.001 ~ "***"))
```

#### Mean and median species JCD across grains

-   Table

```{r}
#| label: sp-mean-sac-grain
#| message: false
#| warning: false

### Median join count across grains plot (species distributions)
# Data preparation (table)
mean_grain_df <- data_species %>% 
  filter(!is.na(jcDifference), 
         is.finite(jcDifference)) %>% 
  group_by(datasetID, samplingPeriodID, startYear, cellSide) %>%
  summarise(meanJCD = mean(jcDifference),
            medianJCD = median(jcDifference), 
            .groups = 'drop') %>% 
  select(datasetID, samplingPeriodID, startYear, cellSide, meanJCD, medianJCD) %>%
  mutate(datasetID = factor(datasetID, levels = c("Czechia", "Europe", "New York", "Japan"))) %>% 
  ungroup()
```

-   Plot

```{r}
#| label: plot-sp-mean-sac-grain
#| message: false
#| warning: false

# Plot
ps2_means <- ggplot(data = mean_grain_df,
                           aes(x = cellSide, 
                               y = meanJCD,
                               color = datasetID, 
                               group = interaction(datasetID, samplingPeriodID),
                               shape = factor(samplingPeriodID),
                               linetype = factor(samplingPeriodID))) + 
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_line(size = 1, show.legend = TRUE) +  
  geom_point(size = 2, show.legend = TRUE) +  
  scale_x_log10() +
  scale_color_manual(values = region_col) +  
  scale_shape_manual(values = c(16, 17, 18), name = "Atlas replication") +  
  scale_linetype_manual(values = c("solid", "dashed", "dotted"), name = "Atlas replication") +
  labs(x = "Cell side (km)",
       y = "JCD mean") +
  ylim(min(data_species$jcDifference, na.rm = T) - 0.01,
       max(data_species$jcDifference, na.rm = T) + 0.01) +
  theme_classic() +
  theme(
    legend.title = element_text(size = 12),  
    legend.text = element_text(size = 12),  
    legend.justification = c(1, 1),  
    plot.margin = margin(t = 5, r = 15, b = 5, l = 5),  
    legend.position = c(0.99, 1),  
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.text.y = element_text(size = 10),
    legend.spacing.y = unit(0.15, "cm"),  
    legend.key.size = unit(0.5, "cm"),  
    legend.key.width = unit(1.5, "cm")  # More space for better visibility
  ) +
  guides(
    shape = guide_legend(title = "Atlas replication",
                         override.aes = list(size = 2)),  # Bigger symbols
    linetype = guide_legend(title = "Atlas replication",
                            override.aes = list(size = 1)),  # Thicker legend lines
    color = guide_legend(title = "Atlas location")  
  )


# Plot
print(ps2_means)
```

-   Saving both

```{r}
#| label: save-sp-mean-sac-grain
#| message: false
#| warning: false

# Save median JCD grain table
write.csv(mean_grain_df, file.path(paths$output_path, "mean_grain_sac_sp.csv"))

# Save plot
ggsave(plot = ps2_means, dpi = 300, width = 5, height = 6,
       file.path(paths$figure_path,
                 "ps2_sac_grain_mean.png"))
```

### 4. Joining species richness and distribution plots

```{r}
#| label: join-sac-grain-plot
#| message: false
#| warning: false

# Arranging all the plots into one (Richness Moran's I and Species Join count)
plot_grain_fig <- grid.arrange(
  arrangeGrob(grain_p5rich_list$Czechia, 
              grain_p5sp_list$Czechia, 
              widths = c(1, 3)),
  arrangeGrob(grain_p5rich_list$Europe, 
              grain_p5sp_list$Europe, 
              widths = c(1, 3)),
  arrangeGrob(grain_p5rich_list$`New York`, 
              grain_p5sp_list$`New York`, 
              widths = c(1, 3)),
  arrangeGrob(grain_p5rich_list$Japan, 
              grain_p5sp_list$Japan, 
              widths = c(1, 3)),
  nrow = 4
)
```

-   Saving plot

```{r}
#| label: save-sac-grain-plot
#| message: false
#| warning: false

# Saving
ggsave(plot = plot_grain_fig, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "p5_sac_grain.png"))
```
