---
title: "10_analyses_traits"
author: "Carmen Soria"
format: html
editor: visual
---

# SAC and species traits

Here I am going to analyse if different categorical groups have different SAC.

```{r}
#| label: libraries
#| message: false
#| warning: false


pacman::p_load(tidyverse, sf, viridis, gridExtra, grid, cowplot, here, broom, rstatix, readxl, ggpubr)
```

### Directories

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Variables

```{r}
# Palette log ratio quadrants
quadrant_col <- c("Q1" = "#FEE090","Q2" = "#FC8D59", "Q3" = "#D73027", 
                  "Q4" = "#87cfc7", "Q5" = "#2e78a0", "Q6" = "#09177e",
                  "No change" = "#4A4A4A", 
                  "Pos. Infinite" = "#F0C24E", "Neg. Infinite" = "#B28300",
                  "Change JC" = "#8A5AC7", "Change OCC" = "#C8A5F0",
                  "Equal" = "#ff499e")

# Palette atlas regions
region_col <-  c("Czechia" = "#4477AA", "Europe" = "#EE6677", "New York" = "#228833", "Japan" = "#CCBB44")
```

### Load data

```{r}
#| label: load-data
#| message: false
#| warning: false

# Trait data
trait_data <- read_excel(file.path(paths$trait_path, "AVONET", "AVONET Supplementary dataset 1.xlsx"), sheet = 2) # Avonet1_birdlife sheet

# SAC species data
data_species <- read.csv(file.path(paths$output_path, "sac_species_table.csv")) %>%
  dplyr::select(-X) %>% 
  mutate(
    datasetID = factor(datasetID, 
                       levels = c("Czechia", "Europe", "New York", "Japan")),
    cellSide = case_when(
      datasetID == "Czechia" ~ scalingID * 11, # Cell side = 11 km
      datasetID == "Europe" ~ scalingID * 50, # Cell side = 50 km
      datasetID == "New York" ~ scalingID * 5, # Cell side = 5 km
      datasetID == "Japan" ~ scalingID * 20, # Cell side = 50 km
      )
    )

# Data JC and OCC change
sp_jcocc_data <- read.csv(file.path(paths$output_path, "logratio_jcocc.csv")) %>% 
  dplyr::select(-X) %>% 
  mutate(datasetID = factor(datasetID, 
                       levels = c("Czechia", 
                                  "Europe", 
                                  "New York", 
                                  "Japan")))


# Dataset IDs
datasetIDs <- unique(sp_jcocc_data$datasetID)
```

## Data preparation

### Trait and BBA taxonomic match

```{r}
tax_df <- data_species %>% 
  left_join(trait_data %>% select(Species1, Trophic.Level),
            by = c("scientificName" = "Species1")) %>% 
  select(scientificName, Trophic.Level) %>% 
  mutate(avonetName = ifelse(!is.na(Trophic.Level), scientificName, NA)) %>%
  select(-Trophic.Level) %>% 
  unique() %>% 
  mutate(avonetName = case_when(
    is.na(avonetName) & word(scientificName, 1) == "Curruca" ~ paste("Sylvia", word(scientificName, 2)),
    is.na(avonetName) & word(scientificName, 1) == "Helopsaltes" ~  paste("Locustella", word(scientificName, 2)),
    scientificName == "Tetrastes bonasia" ~ "Bonasa bonasia",
    scientificName == "Cuculus optatus" ~ "Cuculus saturatus",
    scientificName == "Luscinia svecica" ~ "Cyanecula svecica",
    scientificName == "Canachites canadensis" ~ "Falcipennis canadensis",
    scientificName == "Pterorhinus perspicillatus" ~ "Garrulax perspicillatus",
    scientificName == "Grus canadensis" ~ "Antigone canadensis",
    scientificName == "Padda oryzivora" ~ "Lonchura oryzivora",
    scientificName == "Corthylio calendula" ~ "Regulus calendula",
    .default = avonetName
  )) %>% 
  mutate(avonetName = ifelse(avonetName == "Locustella fasciolatus", "Locustella fasciolata", avonetName))
```

### JC Difference temporal slopes + traits

Process: filter only to keep rows that belong to the original scale and that have a JCD value. Filter for species that are present in all periods and run a linear regression. Join with traits.

```{r}
# JCD data
jcd_data <- data_species %>% 
  filter(scalingID == 1 & !is.na(jcDifference)) %>% 
  select(datasetID, startYear, scientificName, jcDifference) %>% 
  group_by(datasetID, scientificName) %>% 
  mutate(n_periods = n()) %>% 
  ungroup() %>% 
  group_by(datasetID) %>% 
  filter(n_periods == max(n_periods)) %>% 
  ungroup() %>% 
  group_by(datasetID, scientificName) %>% 
  arrange(startYear, .by_group = TRUE) %>% 
  group_modify(~ {
    mod <- lm(jcDifference ~ startYear, data = .x)
    slope_data <- tidy(mod) %>% filter(term == "startYear")
    r2_data <- glance(mod) %>% select(r.squared)
    bind_cols(slope_data, r2_data) 
  }) %>% 
  ungroup() %>% 
  rename(JCD_slope = estimate,
         se_slope = std.error,
         r2 = r.squared) %>% 
  left_join(tax_df, by = "scientificName") %>% 
  left_join(trait_data %>% 
            select(Species1, `Hand-Wing.Index`, Trophic.Level, Habitat.Density), 
            by = c("avonetName" = "Species1")) %>% 
  mutate(Habitat.Density = as.factor(Habitat.Density),
         Trophic.Level = as.factor(Trophic.Level)) %>% 
  rename(habitatDensity = Habitat.Density,
         handWingIndex = `Hand-Wing.Index`,
         trophicLevel = Trophic.Level)
```

### Join log ratio observed JC Statistic with trait data

```{r}
# Log ratio data
jcocc_data <- sp_jcocc_data %>% 
  filter(scalingID == 1) %>% 
  left_join(tax_df, by = "scientificName") %>% 
  left_join(trait_data %>% 
              select(Species1, `Hand-Wing.Index`, Trophic.Level, Habitat.Density), by = c("avonetName" = "Species1")) %>% 
  mutate(Habitat.Density = as.factor(Habitat.Density),
         Trophic.Level = as.factor(Trophic.Level)) %>% 
  rename(habitatDensity = Habitat.Density,
         handWingIndex = `Hand-Wing.Index`,
         trophicLevel = Trophic.Level)
```

### Residuals of logratio JC Statistic respective logratio expJC \~ logratio OCC

```{r}
resid_lm <- jcocc_data %>% 
  group_by(datasetID) %>% 
  group_modify( ~ {
    noinf <- filter(., is.finite(logratioJCExp) & is.finite(logratioJCStat))
    lm_exp <- lm(logratioJCExp ~ logratioOcc, data = noinf)
    mutate(noinf,
           predJCExp = predict(lm_exp, newdata = noinf),
           resid_obs_exp = logratioJCStat - predJCExp)
  }) %>% 
  ungroup()
```

## Trait 1: Habitat density

### JCD temporal slopes

-   Statistics

```{r}
# Check if the distributions of jcd are normal within each group
habjcd_norm <- jcd_data %>% 
  group_by(datasetID, habitatDensity) %>% 
  summarise(shapiro_p = shapiro.test(JCD_slope)$p.value,
            n = n(), .groups = "drop") %>% 
  mutate(normal_distr = ifelse(shapiro_p < 0.05, "no", "yes"))

table(habjcd_norm$normal_distr) # Most don't have a normal distribution

# Kruskal-Wallis test per dataset
kruskal_habjcd <- jcd_data %>%
  group_by(datasetID) %>%
  kruskal_test(JCD_slope ~ habitatDensity) %>%
  ungroup() 

kruskal_habjcd # None of them are significant


# Pairwise comparisons (Dunn test)
pairwise_habjcd <- jcd_data %>%
  group_by(datasetID) %>%
  dunn_test(JCD_slope ~ habitatDensity, p.adjust.method = "bonferroni") %>%
  ungroup()
pairwise_habjcd # None of them are significant
```

-   Plot function

```{r}
plot.ps11 <- function(dataset_name, data) {
  
  dataset_df <- data %>% 
    filter(datasetID == dataset_name)
  
  # Count species per habitat × cellSide
  count_df <- dataset_df %>% 
    group_by(habitatDensity) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  # One common ypos across all facets
  ypos <- max(dataset_df$JCD_slope, na.rm = TRUE) * 1.15
  
  # Plot
  ggplot(dataset_df, aes(x = habitatDensity, y = JCD_slope)) +
    geom_jitter(aes(colour = factor(habitatDensity)), size = 2, alpha = 0.5, 
                width = 0.15, height = 0) +
    geom_boxplot(fill = NA, colour = "black", width = 0.5, outlier.shape = NA) +
    scale_color_viridis_d(name = "Habitat density",
                          labels = c("Dense", "Semi-open", "Open")) +
    geom_text(
      data = count_df,
      aes(x = habitatDensity, y = ypos, label = paste0("N=", n_sp)),
      inherit.aes = FALSE, size = 2.5) +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_x_discrete(labels = c("Dense", "Semi-open", "Open")) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
    theme(
      legend.position = "right",
      axis.title.x = element_blank(),
      axis.text.x = element_blank(), 
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
      axis.ticks.x = element_blank(),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
    )
}
```

-   Plotting

```{r}
# Build plots
ps11_list <- purrr::map(datasetIDs, ~ plot.ps11(.x, jcd_data))
names(ps11_list) <- datasetIDs

# Get legend
ps11_legend <- cowplot::get_legend(ps11_list$Czechia + theme(legend.position = "right"))

# Remove legends from plots
ps11_noleg <- purrr::map(ps11_list, ~ .x + theme(legend.position = "none"))

# Stack plots with one common legend (left)
ps11_all <- cowplot::plot_grid(
  cowplot::plot_grid(plotlist = ps11_noleg, ncol = 2),
  ps11_legend,
  rel_widths = c(1, 0.245)
)
ps11_all <- ggdraw() +
  draw_plot(ps11_all) +
  draw_label("JCD slope", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps11_all
```

-   Save plots

```{r}
ggsave(plot = ps11_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps11_habjcd.png"))
```

### Residuals jcc occ

-   Statistics

```{r}
# Check if the distributions of jcd are normal within each group
habres_norm <- resid_lm %>% 
  group_by(datasetID, habitatDensity) %>% 
  summarise(shapiro_p = shapiro.test(resid_obs_exp)$p.value,
            n = n(), .groups = "drop") %>% 
  mutate(normal_distr = ifelse(shapiro_p < 0.05, "no", "yes"))

table(habres_norm$normal_distr) # None a normal distribution

# Kruskal-Wallis test per dataset
kruskal_habres <- resid_lm %>%
  group_by(datasetID) %>%
  kruskal_test(resid_obs_exp ~ habitatDensity) %>%
  ungroup() 

kruskal_habres

# Pairwise comparisons (Dunn test)
pairwise_habres <- resid_lm %>%
  group_by(datasetID) %>%
  dunn_test(resid_obs_exp ~ habitatDensity, p.adjust.method = "bonferroni") %>%
  ungroup()

pairwise_habres
```

-   Plots

```{r}
plot.ps12 <- function(dataset_name, data, response) {
  dataset_df <- data %>% 
    filter(datasetID == dataset_name)
  
  # Count species per habitat × cellSide
  count_df <- dataset_df %>% 
    group_by(habitatDensity) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  # One common ypos across all facets
  ypos <- max(dataset_df[[response]], na.rm = TRUE) * 1.25
  
  ggplot(dataset_df, aes(x = habitatDensity, y = .data[[response]])) +
    geom_jitter(aes(colour = factor(quadrant)), size = 2, alpha = 0.7, 
                width = 0.15, height = 0) +
    geom_boxplot(fill = NA, colour = "black", width = 0.4, outlier.shape = NA) +
    scale_color_manual(values = quadrant_col, name = "Quadrant") +
    geom_text(
      data = count_df,
      aes(x = habitatDensity, label = paste0("N=", n_sp)),  
      y = ypos,                                 
      inherit.aes = FALSE,
      size = 3
    ) +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_x_discrete(labels = c("Dense", "Semi-open", "Open")) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme(
      legend.position = "right",
      legend.text = element_text(size = 8),
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 8, margin = margin(t = 2)),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
    )
}
```

-   Get plots

```{r}
# Build plots
ps12_list <- purrr::map(datasetIDs, ~ plot.ps12(.x, resid_lm, "resid_obs_exp"))
names(ps12_list) <- datasetIDs

# Extract legend (now it will exist)
ps12_legend <- cowplot::get_legend(ps12_list$Japan + theme(legend.position = "right"))

# Remove legends from plots
ps12_nolegend <- purrr::map(ps12_list, ~ .x + theme(legend.position = "none"))

# Stack plots with one common legend on the left
ps12_all <- cowplot::plot_grid(
  cowplot::plot_grid(plotlist = ps12_nolegend, ncol = 2),
  ps12_legend,
  rel_widths = c(1, 0.24)
)
ps12_all <- ggdraw() +
  draw_plot(ps12_all) +
  draw_label("residuals JC-OCC", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps12_all
```

-   Save plot

```{r}
ggsave(plot = ps12_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps12_habres.png"))
```

## Trait 2: Trophic level

### JCD temporal slopes

-   Statistics

```{r}
# Check if the distributions of the residuals are normal within each group
trojcd_norm <- jcd_data %>% 
  filter(trophicLevel != "Scavenger") %>% # Only 3 in Europe and 1 in NY
  group_by(datasetID, trophicLevel) %>% 
  summarise(shapiro_p = shapiro.test(JCD_slope)$p.value,
            n = n(), .groups = "drop") %>% 
  mutate(normal_distr = ifelse(shapiro_p < 0.05, "no", "yes"))

table(trojcd_norm$normal_distr) # Some yes, most no

# Perform Kruskal-Wallis test per dataset 
kruskal_trojcd <- jcd_data %>%
  group_by(datasetID) %>%
  kruskal_test(JCD_slope ~ trophicLevel) %>%
  ungroup() 

kruskal_trojcd # None of them are significant


# Pairwise comparisons (Dunn test)
pairwise_trojcd <- jcd_data %>%
  group_by(datasetID) %>%
  dunn_test(JCD_slope ~ trophicLevel, p.adjust.method = "bonferroni") %>%
  ungroup()
pairwise_trojcd # None of them are significant
```

-   Plot function

```{r}
plot.ps13 <- function(dataset_name, data) {
  
  dataset_df <- data %>% 
    filter(datasetID == dataset_name) %>% 
    filter(trophicLevel != "Scavenger") 
  
  # Count species per trophic level
  count_df <- dataset_df %>% 
    group_by(trophicLevel) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  # One common ypos across all facets
  ypos <- max(dataset_df$JCD_slope, na.rm = TRUE) * 1.25
  
  ggplot(dataset_df, aes(x = trophicLevel, y = JCD_slope)) +
    geom_jitter(aes(colour = factor(trophicLevel)), size = 2, alpha = 0.5, 
                width = 0.15, height = 0) +
    geom_boxplot(fill = NA, colour = "black", width = 0.5, outlier.shape = NA) +
    scale_color_viridis_d(name = "Trophic level",
                          labels = c("Carnivore", "Herbivore", "Omnivore")) +
    geom_text(
      data = count_df,
      aes(x = trophicLevel, label = paste0("N=", n_sp)),   # only mapped aesthetics
      y = ypos, inherit.aes = FALSE, size = 2.5) +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_x_discrete(labels = c("Carnivore", "Herbivore", "Omnivore")) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
    theme(
      legend.position = "right",
      axis.title.x = element_blank(),
      axis.text.x = element_blank(), 
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
    )
}
```

-   Plotting

```{r}
# Build plots
ps13_list <- purrr::map(datasetIDs, ~ plot.ps13(.x, jcd_data))
names(ps13_list) <- datasetIDs

# Get legend
ps13_legend <- cowplot::get_legend(ps13_list$Czechia + theme(legend.position = "right"))

# Remove legends from plots
ps13_noleg <- purrr::map(ps13_list, ~ .x + theme(legend.position = "none"))

# Stack plots with one common legend (left)
ps13_all <- cowplot::plot_grid(
  cowplot::plot_grid(plotlist = ps13_noleg, ncol = 2),
  ps13_legend,
  rel_widths = c(1, 0.245)
)
ps13_all <- ggdraw() +
  draw_plot(ps13_all) +
  draw_label("JCD slope", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps13_all
```

-   Saving

```{r}
ggsave(plot = ps13_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps13_trojcd.png"))
```

### Residuals jc occ

-   Stats

```{r}
# Check if the distributions of the residuals are normal within each group
trores_norm <- resid_lm %>% 
  filter(trophicLevel != "Scavenger") %>% # Only 3 in Europe and 1 in NY
  group_by(datasetID, trophicLevel) %>% 
  summarise(shapiro_p = shapiro.test(resid_obs_exp)$p.value,
            n = n(), .groups = "drop") %>% 
  mutate(normal_distr = ifelse(shapiro_p < 0.05, "no", "yes"))

table(trores_norm$normal_distr) # None have a normal distribution

# Perform Kruskal-Wallis test per dataset 
kruskal_trores <- resid_lm %>%
  group_by(datasetID) %>%
  kruskal_test(resid_obs_exp ~ trophicLevel) %>%
  ungroup() 

kruskal_trores # Japan and Europe significant


# Pairwise comparisons (Dunn test)
pairwise_trores <- resid_lm %>%
  group_by(datasetID) %>%
  dunn_test(resid_obs_exp ~ trophicLevel, p.adjust.method = "bonferroni") %>%
  ungroup()
pairwise_trores # None of them are significant
```

-   Plot function

```{r}
plot.ps14 <- function(dataset_name, data = resid_lm) {
  dataset_df <- data %>% 
    filter(datasetID == dataset_name) %>% 
    filter(trophicLevel != "Scavenger") 
  
  # Count species per habitat × cellSide
  count_df <- dataset_df %>% 
    group_by(trophicLevel) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  # One common ypos across all facets
  ypos <- max(dataset_df$resid_obs_exp, na.rm = TRUE) * 1.25
  
  ggplot(dataset_df, aes(x = trophicLevel, y = resid_obs_exp)) +
    geom_jitter(aes(colour = factor(quadrant)), size = 2, alpha = 0.7, 
                width = 0.15, height = 0) +
    geom_boxplot(fill = NA, colour = "black", width = 0.5, outlier.shape = NA) +
    scale_color_manual(values = quadrant_col, name = "Quadrant") +
    geom_text(
      data = count_df,
      aes(x = trophicLevel, label = paste0("N=", n_sp)),   
      y = ypos,                                 
      inherit.aes = FALSE,
      size = 3
    ) +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_x_discrete(labels = c("Carn.", "Herb.", "Omn.")) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme(
      legend.position = "right",
      legend.text = element_text(size = 8),
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 8, margin = margin(t = 2)),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
    )
}
```

-   Plots

```{r}
# Build plots
ps14_list <- purrr::map(datasetIDs, ~ plot.ps14(.x, resid_lm))
names(ps14_list) <- datasetIDs

# Extract legend (now it will exist)
ps14_legend <- cowplot::get_legend(ps14_list$Japan + theme(legend.position = "right"))

# Remove legends from plots
ps14_nolegend <- purrr::map(ps14_list, ~ .x + theme(legend.position = "none"))

# Stack plots with one common legend on the left
ps14_all <- cowplot::plot_grid(
  cowplot::plot_grid(plotlist = ps14_nolegend, ncol = 2),
  ps14_legend,
  rel_widths = c(1, 0.3)
)
ps14_all <- ggdraw() +
  draw_plot(ps14_all) +
  draw_label("residuals JC-OCC", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps14_all
```

-   Saving plot

```{r}
ggsave(plot = ps14_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps14_trores.png"))
```

## Trait 3: Hand wing index

### JCD temporal slopes

-   Statistics: linear models

```{r}
# Linear regression
hwijcd_lm <- jcd_data %>% 
  group_by(datasetID) %>% 
  group_modify( ~ {
    lm_hwi <- lm(JCD_slope ~ handWingIndex, data = .)
    slope_data <- tidy(lm_hwi) %>% filter(term == "handWingIndex")
    r2_data <- glance(lm_hwi) %>% select(r.squared)
    bind_cols(slope_data, r2_data) %>%
      mutate(n_replicates = nrow(.x))
  }) %>% 
  ungroup()

hwijcd_lm
```

-   Plot function

```{r}
plot.ps15 <- function(dataset_name, data = jcd_data) {
  
  dataset_df <- data %>% 
    filter(datasetID == dataset_name)
  
  # Count species per habitat
  count_df <- dataset_df %>% 
    group_by(handWingIndex) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  ggplot(dataset_df, aes(x = handWingIndex, y = JCD_slope)) +
    geom_point(color = region_col[[dataset_name]], size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, colour = "black", size = 0.5) +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme(
      plot.margin = margin(t = 10, r = 10, b = 20, l = 20),
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 9, margin = margin(t = 2)),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold", hjust = 0.5)
    )
}
```

-   Plotting

```{r}
# Build plots
ps15_list <- purrr::map(datasetIDs, ~ plot.ps15(.x, jcd_data))
names(ps15_list) <- datasetIDs

ps15_all <- cowplot::plot_grid(plotlist = ps15_list, ncol = 2)
ps15_all <- ggdraw() +
  draw_plot(ps15_all) +
  draw_label("Hand-wing Index", x = 0.5, y = 0.025, size = 12, vjust = 0) +
  draw_label("JCD slope", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps15_all
```

-   Saving

```{r}
ggsave(plot = ps15_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps15_hwijcd.png"))
```

### Residuals jc occ

-   Statistic

```{r}
# Linear regression
hwires_lm <- resid_lm %>% 
  group_by(datasetID) %>% 
  group_modify( ~ {
    lm_hwi <- lm(resid_obs_exp ~ handWingIndex, data = .)
    slope_data <- tidy(lm_hwi) %>% filter(term == "handWingIndex")
    r2_data <- glance(lm_hwi) %>% select(r.squared)
    bind_cols(slope_data, r2_data) %>%
      mutate(n_replicates = nrow(.x))
  }) %>% 
  ungroup()
```

-   Plot function

```{r}
plot.ps16 <- function(dataset_name, data = resid_lm) {
  
  dataset_df <- data %>% 
    filter(datasetID == dataset_name)
  
  # Count species per habitat
  count_df <- dataset_df %>% 
    group_by(handWingIndex) %>% 
    summarise(n_sp = n_distinct(scientificName), .groups = "drop")
  
  ggplot(dataset_df, aes(x = handWingIndex, y = resid_obs_exp)) +
    geom_point(aes(colour = factor(quadrant)), size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, colour = "black", size = 0.5) +
    scale_color_manual(values = quadrant_col, name = "Quadrant") +
    ggtitle(dataset_name) +
    theme_classic() + 
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    theme(
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 9, margin = margin(t = 2)), 
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9),
      plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
      plot.margin = margin(t = 10, r = 10, b = 20, l = 20)
    )
}
```

-   Plots

```{r}
# Plots
ps16_list  <- purrr::map(datasetIDs, plot.ps16)

# Naming them
ps16_list <- setNames(ps16_list, datasetIDs)

ps16_all <- cowplot::plot_grid(plotlist = ps16_list, ncol = 2)
ps16_all <- ggdraw() +
  draw_plot(ps16_all) +
  draw_label("Hand-wing Index", x = 0.5, y = 0.025, size = 12, vjust = 0) +
  draw_label("residuals JC-OCC", x = 0.02, y = 0.5, angle = 90, size = 12, vjust = 1)
ps16_all
```

-   Save plot

```{r}
ggsave(plot = ps16_all, dpi = 600, width = 5.1, height = 6,
       file.path(paths$figure_path, "ps16_hwires.png"))
```
