---
title: "12_additional_checks"
author: "Carmen Soria"
format: html
editor: visual
---

# Additional checks

### Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

pacman::p_load(tidyverse, sf, here, spdep, mapview)
```

### Directories

```{r}
#| label: directory-variables
#| message: false
#| warning: false

# File with paths
source(file.path(here(), "codes", "00_config_file.R"))
```

### Load data

```{r}
#| label: load-data
#| message: false
#| warning: false

# Species data
species_sac <- read.csv(file.path(paths$output_path, "sac_species_table.csv")) %>%
  dplyr::select(-X)

# Log ratio data
jcocc_df <- read.csv(file.path(paths$output_path, "logratio_jcocc.csv")) %>% 
  select(-X)

# Richness sf
richness_sf <- readRDS(file.path(paths$output_path, "richness_sf.rds"))

# Taxonomy data
taxonomy_df <- readRDS(file.path(paths$input_path, "taxonomy.rds"))

# Original data table extracted from the db
species_data <- readRDS(file.path(paths$input_path, "species_data.rds"))
```

## Temporal changes in occupancy

```{r}
#| label: slopes-occupancy-time
#| message: false
#| warning: false

# Getting the slopes of occupancy against atlas period
slopes_occ_df <- species_sac %>%
  group_by(datasetID, scalingID) %>%
  summarise(
    trendOCC = round(coef(lm(occCells ~ startYear, data = cur_data()))[2], 5),
    p_value = summary(lm(occCells ~ startYear, 
                         data = cur_data()))$coefficients[2,4],
    r_squared = summary(lm(occCells ~ startYear, 
                           data = cur_data()))$r.squared,
    se_trendOCC = summary(lm(occCells ~ startYear, 
                            data = cur_data()))$coefficients[2,2]
  ) %>%
  ungroup() %>%
  select(datasetID, scalingID, trendOCC, se_trendOCC, p_value, r_squared)
```

-   Saving

```{r}
#| label: save-slopes-occupancy-time
#| message: false
#| warning: false

# Slope df (all)
write.csv(slopes_occ_df, file.path(paths$output_path, "slopes_occ_df.csv"))
```

## Histograms of log ratios

### Log ratio SAC

One plot per datasetID and scalingID

```{r}
#| label: histograms-logratios-jc
#| message: false
#| warning: false

# Create histograms for logratioJCStat, faceted by both datasetID and scalingID
plot_jcs1 <- jcocc_df %>%
  ggplot(aes(x = logratioJCStat, fill = factor(scalingID))) +
  geom_histogram(alpha = 0.6, bins = 50) +
  facet_grid(datasetID ~ scalingID) +  # Separate plot for each combination of datasetID and scalingID
  labs(title = "Histogram logratio join count statistic per datasetID and scalingID",
       x = "logratioJCStat", y = "Count", fill = "Scaling ID") +
  theme_minimal()

plot(plot_jcs1)
```

Grouping by scalingID

```{r}
#| label: histograms-logratios-jc-scalingID
#| message: false
#| warning: false


# Create histograms for logratioJCStat, faceted by scalingID
plot_jcs2 <- jcocc_df %>%
  ggplot(aes(x = logratioJCStat, fill = factor(datasetID))) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 50) +
  facet_wrap(~ scalingID) +
  labs(title = "Histogram logratio join count statistic per scalingID",
       x = "logratioJCStat", y = "Count", fill = "Scaling ID") +
  theme_minimal()

plot(plot_jcs2)
```

Saving

```{r}
#| label: save-histograms-logratios-jc
#| message: false
#| warning: false

ggsave(plot = plot_jcs1, dpi = 300,
       file.path(paths$figure_path,
                 "hist_logratio_sac1.png"))

ggsave(plot = plot_jcs2, dpi = 300,
       file.path(paths$figure_path,
                 "hist_logratio_sac2.png"))
```

### Log ratio Occupancy

One plot per dataset and scalingID

```{r}
#| label: histograms-logratios-occ
#| message: false
#| warning: false

# Create histograms for logratio occupancy, faceted by both datasetID and scalingID
plot_occ1 <- jcocc_df %>%
  ggplot(aes(x = logratioOcc, fill = factor(scalingID))) +
  geom_histogram(alpha = 0.6, bins = 50) +
  facet_grid(datasetID ~ scalingID) +  # Create a separate plot for each combination of datasetID and scalingID
  labs(title = "Histogram logratio occupancy per datasetID and scalingID",
       x = "logratioOcc", y = "Count", fill = "Scaling ID") +
  theme_minimal()

plot(plot_occ1)
```

Grouping by scalingID

```{r}
#| label: histograms-logratios-occ-scalingid
#| message: false
#| warning: false


plot_occ2 <- jcocc_df %>% 
  ggplot(aes(x = logratioOcc, fill = factor(datasetID))) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 50) +
  facet_grid(~scalingID) +
  labs(title = "Histogram logratio occupancy per dataset",
       x = "logratioOcc", y = "Count", fill = "Scaling ID") +
  theme_minimal()

plot(plot_occ2)
```

Saving

```{r}
#| label: save-histograms-logratios-occ
#| message: false
#| warning: false

ggsave(plot = plot_occ1, dpi = 300, 
       file.path(paths$figure_path,
                 "hist_logratio_occ1.png"))

ggsave(plot = plot_occ2, dpi = 300, 
       file.path(paths$figure_path,
                 "hist_logratio_occ2.png"))
```

Removing plots

```{r}
#| label: remove-plots
#| message: false
#| warning: false

rm(list=ls(pattern="^plot_"))
```

## Number of species for which we could obtain SAC

-   Total number of species per region, period and grain size

```{r}
#| label: species-sac-number
#| message: false
#| warning: false

sp_sac <- species_sac %>% 
  select(datasetID, samplingPeriodID, scalingID, scientificName, jcStatistic) %>% 
  group_by(datasetID, samplingPeriodID, scalingID) %>% 
  summarise(nSpSac = n_distinct(scientificName[!is.na(jcStatistic)]),
            totalSp = n_distinct(scientificName),
            percSpSac = (nSpSac / totalSp) * 100, .groups = "drop")
```

-   Number of species for which we were able to calculate the log ratios

```{r}
#| label: species-logratio-number
#| message: false
#| warning: false

sp_logratio <- jcocc_df %>% 
  select(datasetID, scalingID, scientificName, logratioJCStat) %>% 
  group_by(datasetID, scalingID) %>% 
  summarise(nSpSac = n_distinct(scientificName))
```

-   The number of species for which we could calculate the log ratio should be species that have appeared in at least two periods of each atlas for which we could calculate the join count

```{r}
#| label: species-logratio-number-check
#| message: false
#| warning: false

sp_twice <- species_sac %>% 
  select(datasetID, samplingPeriodID, scalingID, scientificName, jcStatistic) %>% 
  filter(!is.na(jcStatistic)) %>% 
  group_by(datasetID, scalingID, scientificName) %>% 
  summarise(nPeriods = n()) %>% 
  ungroup() %>% 
  filter(nPeriods >= 2) %>% 
  group_by(datasetID, scalingID) %>% 
  summarise(nSp = n())


# Are they the same?
identical(sp_logratio$nSpSac, sp_twice$nSp)
```

## Number of cells and neighbors

-   Total number of neighbors per cell

```{r}
#| label: neighbor-cells-number
#| message: false
#| warning: false

sf_use_s2(FALSE)

# Calculate total number of neighbors
nb_tbl <- richness_sf %>% 
  select(datasetID, scalingID, siteID, geometry) %>% 
  unique() %>% 
  group_by(datasetID, scalingID) %>% 
  group_modify(~ {
    
    group_data <- .x
    nb_group <- poly2nb(group_data, queen = T)
    
    group_data$nNeighbors <- card(nb_group)
    
    group_data
  }) %>% 
  ungroup() %>% 
  select(-geometry)

nb_tbl1 <- nb_tbl %>% 
  group_by(datasetID, scalingID, nNeighbors) %>% 
  summarise(countNeighbors = n(), .groups = "drop")
```

-   Number of edge cells per dataset and grain size

```{r}
#| label: edge-cells-number
#| message: false
#| warning: false

nb_tbl2 <- nb_tbl1 %>% 
  filter(nNeighbors < 8 | nNeighbors == 9) %>% # Some EBBA cells are joined and can have a max of 10 neighbors
  group_by(datasetID, scalingID) %>% 
  summarise(nEdgeCells = sum(countNeighbors))
```

-   Joining with total number of cells

```{r}
#| label: neighbor-edge-cells-table
#| message: false
#| warning: false

# total number of cells
ncells_tbl <- richness_sf %>% 
  st_drop_geometry() %>% 
  select(datasetID, scalingID, siteID) %>% 
  unique() %>% 
  group_by(datasetID, scalingID) %>% 
  summarise(nCells = length(siteID))


# Joining
out <- nb_tbl2 %>% 
  full_join(ncells_tbl) %>% 
  mutate(proportion = nEdgeCells / nCells)
```

-   Saving

```{r}
#| label: save-neighbor-edge-cells-table
#| message: false
#| warning: false

write.csv(out, file.path(paths$output_path, "proportion_edge_cells.csv"))
```

## Which species are winners and losers?

### Joining with taxonomy

```{r}
#| label: add-taxonomy-logratios
#| message: false
#| warning: false

jcocc_df1 <- jcocc_df %>% 
  left_join(taxonomy_df, by = "scientificName") %>% 
  select(datasetID, scalingID, cellSide, order, family, genus, scientificName,
         firstJCStat, lastJCStat, logratioJCStat,
         firstOcc, lastOcc, logratioOcc, quadrant)
```

Winners = species that are increasing in occupancy.

```{r}
#| label: winners-logratio-occupancy
#| message: false
#| warning: false

occ_inc <- jcocc_df1 %>% 
  filter(logratioOcc > 0) %>% 
  group_by(datasetID, scalingID, cellSide, family) %>% 
  summarise(nFamily = n())
```

Losers = species that are declining in occupancy.

```{r}
#| label: losers-logratio-occupancy
#| message: false
#| warning: false

occ_dec <- jcocc_df1 %>% 
  filter(logratioOcc < 0) %>% 
  group_by(datasetID, scalingID, cellSide, family) %>% 
  summarise(nFamily = n())
```
